---
title: "Fitted-model labels with 'ggpmisc' and 'plotly' or 'ggiraph'"
subtitle: "Interactive ggplots with model equations, _R_<sup>2</sup>, _F_, _P_, etc."
author: "Pedro J. Aphalo"
date: "2023-06-25"
date-modified: "2026-06-05"
keywords: [ggplot2 pkg, ggpmisc pkg, plotly pkg, ggiraph pkg, plot annotations, model equations]
categories: [plotting, data visualization, ggpp, ggpmisc, ggplot2]
abstract: |
  Example R code for _interactive_ plots based on package [ggplot2](https://ggplot2.tidyverse.org/) using geometries defined in package [ggpp](https://docs.r4photobiology.info/ggpp) and statistics from package [ggpmisc](https://docs.r4photobiology.info/ggpmisc) together with package [plotly](https://plotly.com/ggplot2/). This page is a draft and currently contains a single plot example.
editor: 
  markdown: 
    wrap: 72
format:
  html: 
    code-link: true
    code-tools: true
    html-math-method: mathjax
---

::: callout-warning
At the moment support for math typesetting in interactive plots is very limited in R. 'plotly', at least in older versions, did support rendering of $\LaTeX$-encoded labels with 'mathjax'. Based on issues at GitHub and the very sparse documentation on the subject, this no longer can be expected to work out of the box, if at all. At least the examples in the 'plotly' documentation no longer render as expected.
:::

::: callout-tip
In this page code chunks are "folded" so as to decrease the clutter when
searching for examples. Above each plot you will find a small triangle
followed by "Code". Clicking on the triangle "unfolds" the code chunk
making visible the R code used to produce the plot. Except for the
loading of packages shown in section **Preliminaries** code examples are
in most cases self contained. When they are not, this is indicated by a
comment.

For simplicity, whenever possible I use base R functions instead of
contributed R packages. For those packages used only in specific
examples I use colon notation to indicate the 'package'.

All "words" defined in base R or in extension packages are linked to the
corresponding HTML-rendered help pages.

The code in the chunks can be copied by clicking on the top right
corner, where an icon appears when the mouse cursor hovers over the code
listing.
:::

::: callout-important
One needs to always check that annotations do not occlude anything
significant, such as observations in the base plot. This needs special
care when using annotations together with batch plotting. Either ensure
that the scale limits of the base plot are expanded to avoid overlap or
that the layer with the equations is the lowest one, i.e., added to the
plot first.
:::

# Preliminaries

We first load the packages we will use.

When package 'ggpmisc' is loaded and attached, packages 'ggpp' and
'ggplot2' are also attached. The only function from 'ggplot2' that is
redefined by 'ggpp' is `annotate()`, which remains backwards compatible
with 'ggplot2'.

```{r, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', dev = "svg",
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)
```

```{r, message=FALSE}
#| code-fold: false
library(ggplot2)
library(ggpmisc)
library(plotly)
theme_set(theme_bw())
```

Artificial data.

```{r}
set.seed(1)
df <- data.frame(x = c(1:100))
df$y <- 2 + 3 * df$x + rnorm(100, sd = 40)
df$idx <- paste("id", df$x, sep = "-")
```

# Using 'plotly'

R package 'plotly' can convert a regular ggplot into an interactive plot that
can be embedded in HTML output. Thus the usual Grammar of Graphics familiar 
to 'ggplot2' users can be used to create interactive plots. The disadvantage is that only a rather small subset of existing geoms are supported. The reason is that 'plotly' is implemented as a parsing engine that recognizes only known
'ggplot2' layer functions. 

The geometries defined in R package 'ggpp', used by default in several of the
stats in 'ggpmisc' are not supported. These stats can still be used together with geoms from 'ggplot2' with some limitations. The main one of these is that 'plotly' does not support `plotmath` expressions. The stats from 'ggpmisc' can generate text strings mapped to the `label` aesthetic encoded as plain text but the resulting annotations are not as nice as when rendering `plotmath` expressions using R graphic devices.

Recent versions of 'quarto' do support this approach. 

```{r}
#| fig-cap: Works with Quarto >= 1.4.0
my.formula <- y ~ x
p1 <- ggplot(data = df, aes(x = x, y = y)) +
  stat_poly_line(method = "lm", 
                 se=FALSE, 
                 color="black", 
                 formula = my.formula) +
  stat_poly_eq(geom = "text",
               output.type = "text",
               formula = my.formula,
               mapping = use_label(c("eq", "R2"), sep = ",  "),
               label.x = 20) +
  geom_point()

ggplotly(p1)
```

'ggpmisc' (>= 0.7.0) fully supports encoding labels using $\LaTeX$ markup, potentially opening and additional route to nicely rendered fitted model equations. However, this feature seems to be broken in 'plotly' as the example in the help page `plotly::TeX()` also fails to render correctly. Has $\LaTeX$ support with 'mathjax' been removed? or is this a setup problem?

```{r}
#| fig-cap: Does not work with Quarto 1.9.x!
my.formula <- y ~ x
p2 <- ggplot(data = df, aes(x = x, y = y)) +
  stat_poly_line(method = "lm", 
                 se=FALSE, 
                 color="black", 
                 formula = my.formula) +
  stat_poly_eq(geom = "text",
               output.type = "latex",
               formula = my.formula,
               mapping = aes(label = after_stat(TeX(eq.label))),
               label.x = 20) +
  geom_point()

ggplotly(p2)
```

# Using 'ggiraph'

```{r}
detach(package:plotly)
library(ggiraph)
```

R package 'ggiraph' seems not to render layers with data computed by stats, preventing both model equations and the prediction line from being rendered. Although this example uses stats from 'ggpmisc' the example in the Appendix that relies purely on 'ggplot2' fails in the same way.

```{r}
#| fig-cap: Equation and fitted model line are not displayed!
my.formula <- y ~ x
p3 <- ggplot(data = df, aes(x = x, y = y,
                           tooltip = round(y, 0), data_id = idx)) +
  stat_poly_line(method = "lm", 
                 se = FALSE, 
                 color="black", 
                 formula = my.formula) +
  stat_poly_eq(formula = my.formula,
               mapping = use_label(c("eq", "R2"), sep = ",  "),
               label.x = 20) +
  geom_point_interactive()

girafe(ggobj = p3)
```

## Test

Example not using 'ggpmisc' also fails.

```{r}
detach(package:ggpmisc)
```

```{r}
#| fig-cap: Fitted model line and confidence band are not displayed!
p4 <- ggplot(data = df, aes(x = x, y = y,
                           tooltip = round(y, 0), data_id = idx)) +
  stat_smooth(method = "lm", 
                   formula =  y ~ x) +
  geom_point_interactive()

girafe(ggobj = p4)
```

::: callout-tip
The possibilities remain rather limted at the moment for the creation of interactive plots with 'ggpmisc'.
:::
