---
title: "ggplot fitted-model equations with 'ggpmisc'"
subtitle: "Polynomials, major axis, quantile and non-linear regression"
author: "Pedro J. Aphalo"
date: "2023-02-27"
date-modified: "2023-07-30"
keywords: [ggplot2 pkg, ggpp pkg, ggpmisc pkg, data labels, plot annotations, model equations]
categories: [Plotting examples]
editor: 
  markdown: 
    wrap: 72
code-fold: true
format:
  html: 
    code-link: true
    code-tools: true
---

::: callout-tip
In this page code chunks are "folded" so as to decrease the clutter when
searching for examples. Above each plot you will find one or more "folded" code
chuncks signalled by a small triangle followed by "Code". Clicking on the
triangle "unfolds" the code chunk making visible the R code used to produce the
plot.

The code in the chunks can be copied by clicking on the top right
corner, where an icon appears when the mouse cursor hovers over the code
listing.

The `</> Code` drop down menu to the right of the page title makes it possible to unfold all code chunks and to view the Quarto source of the whole web page.

Names of functions and other R objects are linked to the corresponding on-line help pages. The names of R extension packages are linked to their documentation web sites when available.
:::

## Introduction

Here you will find examples of ggplots with insets. The insets have been
added to the plots with geometries defined in package [ggpp](https://docs.r4photobiology.info/ggpp/) and
statistics from package [ggpmisc](https://docs.r4photobiology.info/ggpmisc/) both available at CRAN.

::: callout-important
One needs to always check that annotations do not occlude anything
significant, such as observations in the base plot. This needs special
care when using annotations together with batch plotting. Either ensure
that the scale limits of the base plot are expanded to avoid overlap or
that the layer with the equations is the lowest one, i.e., added to the
plot first.
:::

## Data labels and plot annotations

Data labels add textual information directly related to individual data
points (shown as glyphs). Text position in this case is dependent on the
scales used to represent data points. Text is usually displaced so that
it does not occlude the glyph representing the data point and when the
link to the data point is unclear, this link is signalled with a line
segment or arrow. Data labels are distinct from annotations in that they
contribute directly to the representation of data on a plot or map.

Annotations differ from data labels, in that their position is decoupled
from their meaning. Insets can be thought as larger, but still
self-contained annotations. In most cases the reading of inset tables
and plots depends only weakly on the plot or map in which they are
included.

In the case of annotations and insets the designer of a data
visualization has the freedom to locate them anywhere, as long as they
do not occlude features used to describe data. I will use the term
annotation irrespective if the "labels" are textual or graphical.

The equations showing parameter estimates from models fit to data are
normally displayed as annotations. However, it is also possible to
consider them data labels "connected" to individual curves representing
the corresponding model fits.

## Annotations showing parameters estimates

Fitted model equations and other related estimates can be useful in
plots as they ensure that graphical representation as a curve and
numerical values for parameters that cannot be read from the curve
itself are displayed as text on the same plot.

::: callout-note
When adding annotations one should be aware that they add clutter to a
plot, and clutter can make it difficult to see the patterns of interest
in the data represented as points or curves.

So, as usual, less is more, include those annotations that are relevant
to the message conveyed by a plot and nothing more.
:::

Fully automating the construction of fitted model equations is far from
trivial for the general case, so I have implemented automation for
specific types of models: polynomials fitted by OLS as linear models,
polynomials fitted by quantile regression and major-axis regression
fits. They are currently available as pairs of statistics with
consistent interfaces, with each pair consisting in a curve-plotting
statistics and an annotations statistics.

As shown below other model formulas can be rather easily assembled and
added as annotations if one is familar with R's expressions as used for
`plotmath`.

## Preliminaries

The code used is shown on-demand above each plot and can be copied. We
first load the packages we will use.

When package [ggpmisc](https://docs.r4photobiology.info/ggpmisc/) is loaded and attached, packages [ggpp](https://docs.r4photobiology.info/ggpp/) and
[ggplot2](https://ggplot2.tidyverse.org/) are also attached. The only function from [ggplot2](https://ggplot2.tidyverse.org/) that is
redefined by [ggpp](https://docs.r4photobiology.info/ggpp/) is `annotate()`, which remains backwards compatible
with [ggplot2](https://ggplot2.tidyverse.org/).

```{r, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', dev = "svg",
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)
```

```{r, message=FALSE}
#| code-fold: false
library(ggpmisc)
library(ggrepel)
library(dplyr)
```

## Linear correlation

```{r}
# generate artifical data for a linear relationship
# two groups with different standard deviation around a slope = 1
set.seed(94321)
x <- (1:100) / 10
yA <- x + rnorm(length(x), sd = 2)
yB <- x + rnorm(length(x), sd = 8)
df1 <- data.frame(x = rep(x, 2),
                  y = c(yA, yB),
                  group = rep(c("A", "B"), rep(length(x), 2L)))
```

The examples in this section make use of `stat_correlation()` from package
[ggpmisc](https://docs.r4photobiology.info/ggpmisc/).

Pearson correlation.

```{r}
ggplot(subset(df1, group == "A"), 
       aes(x, y)) +
  geom_point() +
  stat_correlation()
```

```{r}
ggplot(subset(df1, group == "A"), 
       aes(x, y)) +
  geom_point() +
  stat_correlation(small.r = TRUE)
```

Spearman correlation.

```{r}
ggplot(subset(df1, group == "A"), 
       aes(x, y)) +
  geom_point() +
  stat_correlation(method = "spearman")
```

Kendall correlation.

```{r}
ggplot(subset(df1, group == "A"), 
       aes(x, y)) +
  geom_point() +
  stat_correlation(method = "kendall")
```

Select which labels to show in plot.

```{r}
ggplot(subset(df1, group == "A"), aes(x, y)) +
  geom_point() +
  stat_correlation(mapping = use_label(c("R", "P", "n")),
                   method = "kendall")
```

Select which labels to show in plot.

```{r}
ggplot(subset(df1, group == "A"), aes(x, y)) +
  geom_point() +
  stat_correlation(mapping = use_label(c("R", "t", "P", "n")))
```

Confidence interval for R.

```{r}
ggplot(subset(df1, group == "A"), aes(x, y)) +
  geom_point() +
  stat_correlation(mapping = use_label(c("R", "R.confint")),
                   r.conf.level = 0.99)
```

Grouping supported.

```{r}
ggplot(df1, aes(x, y, color = group)) +
  geom_point() +
  stat_correlation()
```

Facets supported.

```{r}
ggplot(df1, aes(x, y)) +
  geom_point() +
  stat_correlation(npcx = "right", npcy = "bottom") +
  facet_wrap(~group)
```

Highlighting based on estimates, here estimated *R* but it is possible
to use other estimates like *P*-value.

```{r}
ggplot(df1, aes(x, y)) +
  geom_point() +
  stat_correlation(mapping = 
                     aes(color = ifelse(after_stat(cor) > 0.5,
                                        "red", "black"))) +
  scale_color_identity() +
  facet_wrap(~group)
```

### Alternatives

::: callout-tip
Package 'ggpubr' provides alternative approaches to corelation testing
and reporting.
:::

## Polynomials

Polynomials as linear models, but many other linear models exist. The most
common case is linear regression. They are very frequently used, and in the
case of linear regression the role of parameters can be easily recognized.

The examples in this section make use of `stat_poly_line()` and `stat_poly_eq()`
from package [ggpmisc](https://docs.r4photobiology.info/ggpmisc/).

The model corresponding to lineal regression is a first degree polynomial.

```{r}
ggplot(df1, aes(x, y)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(use_label("eq"), 
               label.x = "right", label.y = "bottom") +
  facet_wrap(~group)
```

Most likely the variables mapped onto aesthetics `x` and `y` have different names. If we use a symbol to represent these variables, we may want to use these symbols in the equation.

```{r}
ggplot(df1, aes(x, y)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(use_label("eq"), 
               eq.with.lhs = "italic(t)~`=`~", 
               eq.x.rhs = "~italic(E)",
               label.x = "right", label.y = "bottom") +
  labs(x = expression("Irrandiance, "*italic(E)~~(W~m^{-2})),
       y = expression("Temperature, "*italic(t)~~(degree*"C"))) +
  facet_wrap(~group)
```

```{r}
set.seed(4321)
# generate artificial data for a 3rd degree polynomial
x <- 1:100
y <- (x + x^2 + x^3) + rnorm(length(x), mean = 0, sd = mean(x^3) / 4)
y <- y / max(y)
df2 <- data.frame(x, 
                  y, 
                  group = c("A", "B"), 
                  y2 = y * c(1, 2) + c(0, 0.2),
                  block = c("a", "a", "b", "b"),
                  wt = sqrt(x))
```

$R^2$ is the default label, as its estimate is always available.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(formula = formula)
```

Fitted model equation, available for polynomials with no missing terms.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(mapping = use_label("eq"), formula = formula)
```

$R_\mathrm{adj}^2$ and *P*-value added separately.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(mapping = use_label("adj.R2"), formula = formula) +
  stat_poly_eq(mapping = use_label("P"), label.x = "right", label.y = "bottom", size = 3,
               formula = formula)
```

Fitted model equation and $R_\mathrm{adj}^2$.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(mapping = use_label(c("eq", "adj.R2"), sep = "~~italic(\"with\")~~"),
               formula = formula)
```

Equation with no left-hand-side (lhs).

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(mapping = use_label("eq"),
               eq.with.lhs = FALSE,
               formula = formula)
```

Variable names in the equation set to be meaningful.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(mapping = use_label(c("eq", "R2")),
               eq.with.lhs = "italic(h)~`=`~",
               eq.x.rhs = "~italic(z)",
               formula = formula) +
  labs(x = expression(italic(z)), y = expression(italic(h)))
```

Even Greek charactes and maths can be incorporated when needed.

```{r}
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(df2, aes(x, log10(y + 1e6))) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(mapping = use_label("eq"),
               eq.with.lhs = "plain(log)[10](italic(delta)+10^6)~`=`~",
               eq.x.rhs = "~Omega",
               formula = formula) +
  labs(y = expression(plain(log)[10](italic(delta)+10^6)), 
       x = expression(Omega))
```

::: callout-caution
`stat_poly_eq()` by default assumes that the fitted polynomial has the variable
mapped to the `x` or `y` aesthetic as explanatory variable depending on the
model formula and argument passed to parameter `orientation`. It does not decode
any transformations applied on the fly to `x` or `y`. In addition it does not
function correctly with linear models that are not polynomials.

If a transformation is applied, the user is responsible for replacing as shown
below the $x$ (and/or $y$) symbol in the equation by setting an alternative 
value matching the transformation applied in the model formula.

```{r}
formula <- y ~ I(x^2) # we apply a transformation to x within the formula
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(mapping = use_label("eq"), 
               eq.x.rhs = "~x^2",
               formula = formula)
```

In some cases, like when applying a power as transformation and fitting higher
degree polynomials, one may need to manually build the R expression representing
the equation to simplify it.
:::

The examples shown above also work with groups.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y2, colour = group)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = after_stat(eq.label)), 
               formula = formula, 
               vstep = 0.08) +
  theme_bw()
```

Using colours to indicate the groups to which equations correspond is
not always best. Here we use labels to the left of each equation.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y2, linetype = group, grp.label = group)) +
  geom_point() +
  stat_poly_line(formula = formula, color = "black") +
  stat_poly_eq(aes(label = after_stat(paste("bold(", grp.label, "*':')~~~", 
                                      eq.label, sep = ""))),
               formula = formula)
```

Facets are also supported.

```{r, fig.width = 7, fig.height = 7}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y2)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = after_stat(eq.label)),
               formula = formula) +
  facet_wrap(facets = vars(group), scales = "free_y", ncol = 1)
```

In the stats defined in [ggpmsic](https://docs.r4photobiology.info/ggpmisc/) the fit method function can modify the
model passed as argument because the one used to build the equation is
retrieved from the fitted model object when possible. In this example, a
linear regression is fitted if the slope is significant, but if not the
mean is reported, both graphically and numerically.

```{r, fig.width = 7, fig.height = 9}
# user defined fit method removes the slope if 
# the slope is not significant
poly_or_mean <- function(formula, data, ...) {
   fm <- lm(formula = formula, data = data, ...)
   if (anova(fm)[["Pr(>F)"]][1] > 0.1) {
      lm(formula = y ~ 1, data = data, ...)
   } else {
      fm
   }
}

# we create a plot as usual, but with our method
ggplot(mpg, aes(displ, hwy)) +
   geom_point() +
   stat_poly_line(method = "poly_or_mean") +
   stat_poly_eq(method = poly_or_mean,
   aes(label = after_stat(eq.label)),
   label.x = "right") +
   theme(legend.position = "bottom") +
   facet_wrap(~class, ncol = 2)
```

Faceting works also with free *x* and/or *y* scales in panels.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y2, fill = block)) +
  geom_point(shape = 21, size = 3) +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = after_stat(rr.label)), size = 3,
               geom = "label_npc", alpha = 0.33,
               formula = formula) +
  facet_wrap(~group, scales = "free_y") +
  theme(legend.position = "top")
```

Faceting works also with free *x* and/or *y* scales in panels when the
position of labels is set by a string.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(df2, aes(x, y2, fill = block)) +
  geom_point(shape = 21, size = 3) +
  stat_poly_line(formula = formula) +
  stat_poly_eq(use_label("R2"), size = 3,
               geom = "label_npc", alpha = 0.33,
               formula = formula,
               label.x = "right", label.y = "bottom",
               rr.digits = 3) +
  stat_poly_eq(use_label(c("F", "P")),
               size = 3,
               geom = "label_npc", alpha = 0.33,
               formula = formula) +
  facet_wrap(~group, scales = "free_y") +
  theme(legend.position = "top")
```

Some artificial data with rather high random variation on both _x_ and _y_.

```{r}
set.seed(94321)
a <- (1:100) / 10
x <- a + rnorm(length(x), sd = 1)
y <- a + rnorm(length(x), sd = 2)
df3 <- data.frame(x = x, y = y)
```

Linear regression through the origin.

```{r}
# model with intercept = 0
formula <- y ~ x + 0
ggplot(df3, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(formula = formula,
               mapping = use_label("eq"))
```

Regressions of *y* on *x* and *x* on *y* in the same plot. The lines are
drawn for the range of the explanatory variable, which differs in this
case between the two fits.

```{r}
# the default for formula is y ~ x
ggplot(df3, aes(x, y)) +
  geom_point() +
  stat_poly_line(color = "blue") +
  stat_poly_eq(mapping = use_label(c("R2", "eq")), 
               color = "blue") +
  stat_poly_line(color = "red", 
                 orientation = "y") +
  stat_poly_eq(mapping = use_label(c("R2", "eq")), 
               color = "red", 
               orientation = "y",
               label.y = 0.9)
```

### Alternatives

::: callout-tip
Package 'ggpubr' provides a renamed copy of `stat_poly_eq()` taken from
an old version of [ggpmisc](https://docs.r4photobiology.info/ggpmisc/). The version in 'ggpubr' is much more
limited in its functionality and even contains bugs. The line drawing
code in `stat_smooth()` from [ggplot2](https://ggplot2.tidyverse.org/) was used as a basis for
`stat_poly_line()`, the main difference is that by default it uses
`lm()` as method irrespective of the data size while `stat_smooth()`
switches to using splines for large numbers of observations.
:::

## Major axis regression

If *x* and *y* are both subject to random errors, and none is clearly
the cause of the other, we should use major axis regression instead of
linear regression.

The examples in this section make use of `stat_ma_line()` and `stat_ma_eq()`
from package [ggpmisc](https://docs.r4photobiology.info/ggpmisc/).

```{r}
ggplot(df3, aes(x, y)) +
  geom_point() +
  stat_ma_line() +
  stat_ma_eq(mapping = use_label(c("R2", "eq")))
```

::: callout-tip
No alternatives I know of.
:::

## Quantile regression

Quantile regression is frequently used is some disciplines like
Economics but less frequently in other fields. It is robust and can be
rather easily interpreted based on its similarity to the well known box
plots.

The examples in this section make use of `stat_quant_line()`,
`stat_quant_band()` and `stat_quant_eq()` from package [ggpmisc](https://docs.r4photobiology.info/ggpmisc/).

::: callout-caution
Quantile regression fits are done by numerical approximation, and
frequently generate warnings for non-unique solutions or similar
problems. Their importance needs to be assessed by users. In the
examples below they seem mostly unimportant given the large number of
observations and have been silenced in the output.
:::

A median regression with a band limited by the upper and lower quartile
regressions.

```{r, warning=FALSE}
# the default for formula is y ~ x
ggplot(df3, aes(x, y)) +
  geom_point() +
  stat_quant_band() +
  stat_quant_eq(quantiles = 0.5)
```

Median regression through the origin.

```{r, warning=FALSE}
# model with intercept = 0
formula <- y ~ x + 0
ggplot(df3, aes(x, y)) +
  geom_point() +
  stat_quant_band(formula = formula) +
  stat_quant_eq(formula = formula, quantiles = 0.5)
```

Quantile regressions of _y_ on _x_ and _x_ on _y_.

```{r, warning=FALSE}
# the default for formula is y ~ x
ggplot(df3, aes(x, y)) +
  geom_point() +
  stat_quant_band(color = "blue") +
  stat_quant_eq(quantiles = 0.5, color = "blue") +
  stat_quant_band(color = "red", 
                 orientation = "y") +
  stat_quant_eq(quantiles = 0.5, color = "red", 
               orientation = "y", label.y = 0.9)
```

Fitting a polynomial by quantile regression.

```{r, warning=FALSE}
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_quant_band(formula = y ~ poly(x, 2)) +
  stat_quant_eq(formula = y ~ poly(x, 2), quantiles = 0.5)
```

Two quantiles are by default plotted as lines,

```{r, warning=FALSE}
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_quant_line(formula = y ~ poly(x, 2), quantiles = c(0.05, 0.95)) +
  stat_quant_eq(formula = y ~ poly(x, 2), quantiles = c(0.05, 0.95))
```

A single quantile, here the median, is plotted as a line plus a
confidence band.

```{r, warning=FALSE}
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_quant_line(formula = y ~ poly(x, 2), quantiles = 0.5) +
  stat_quant_eq(formula = y ~ poly(x, 2), quantiles = 0.5)
```

Equations labelled by quantile.

```{r, warning=FALSE}
ggplot(df2, aes(x, y)) +
  geom_point() +
  stat_quant_band(formula = y ~ poly(x, 2), 
                  color = "black", fill = "grey60") +
  stat_quant_eq(aes(label = paste(after_stat(grp.label), "*\": \"*",
                                  after_stat(eq.label), sep = "")),
                formula = y ~ poly(x, 2)) +
  theme_classic()
```

Equations labelled by group and quantile.

```{r, warning=FALSE}
ggplot(df2, aes(x, y, group = group, linetype = group, 
                    shape = group, grp.label = group)) +
  geom_point() +
  stat_quant_line(formula = y ~ poly(x, 2), 
                  quantiles = c(0.1, 0.9), 
                  color = "black") +
  stat_quant_eq(aes(label = paste(after_stat(grp.label), "*\": \"*",
                                  after_stat(eq.label), sep = "")),
                formula = y ~ poly(x, 2), 
                quantiles = c(0.1, 0.9)) +
  theme_classic()
```

### Alternatives

::: callout-tip
Package [ggplot2](https://ggplot2.tidyverse.org/) defines `stat_quantile()` which can be used to plot as
a line some types of quantile regression fits, but is less flexible than
`stat_quant_line()` and `stat_quant_band()`.
:::

## Non-linear models

The examples in previous sections were based on ready assembled labels. Here
we show examples of how to generate labels for equations and other parameter
estimates with an approach applicable to many different types of models.

The examples in this section make use of `stat_fit_tidy()` from package
[ggpmisc](https://docs.r4photobiology.info/ggpmisc/).

We use as example a fit of the Michaelis-Menthen equation of reaction kinetics,
a function that is non-linear in its parameters.

```{r}
#| code-fold: false
# We use methods from package 'broom'
library(broom)
```

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tidy(method = "nls", 
                method.args = list(formula = micmen.formula),
                label.x = "right",
                label.y = "bottom",
                aes(label = paste("V[m]~`=`~", signif(after_stat(Vm_estimate), digits = 3),
                                  "%+-%", signif(after_stat(Vm_se), digits = 2),
                                  "~~~~K~`=`~", signif(after_stat(K_estimate), digits = 3),
                                  "%+-%", signif(after_stat(K_se), digits = 2),
                                  sep = "")),
                parse = TRUE) +
  theme_bw()
```

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tidy(method = "nls", 
                method.args = list(formula = micmen.formula),
                label.x = "center",
                label.y = "bottom",
                vstep = 0, hstep = -0.3,
                aes(label = paste("V~`=`~frac(", signif(after_stat(Vm_estimate), digits = 2), "~C,",
                                  signif(after_stat(K_estimate), digits = 2), "+C)",
                                  sep = "")),
                parse = TRUE) +
  labs(x = "C", y = "V") +
  theme_bw()
```

### Alternatives

::: callout-tip
In all cases it is possible to do the model fitting before creating the plot, assembling
labels in one's own R code and passing them to `ggplot()` as data. The
possibilities are nearly unlimited but implementing them, depending on one's
own familiarity with R and 'ggplot2' can become very time consuming. I have
in part written [ggpp](https://docs.r4photobiology.info/ggpp/) and [ggpmisc](https://docs.r4photobiology.info/ggpmisc/) to avoid as a user having to remind
myself of how to write such code repeatedly. My hope is that others will also
save time and effort by using the packages I have developed.
:::
