---
title: "Plot insets"
author: "Pedro J. Aphalo"
date: today
categories: [Gallery, Plot insets, ggpp pkg]
editor: 
  markdown: 
    wrap: 72
---

## Introduction

The use of insets in plots was rather frequent when plots where drawn
using ink and pen by draughtsmen. With the advent of computer-drawn
plots they became less common. In many situations they are very useful.
In this gallery I give examples of plots with insets created using
packages 'ggplot2' and 'ggpp'.

Insets are not normally used for data exploration, they are in most
cases used in publications and reports, either to provide context or
save space.

::: callout-important
One needs to always check that insets do not occlude anything
significant, such as observations in the base plot. This makes insets
unsuitable for automation, unless the scale limits of the base plot are
expanded to ensure that no overlap is possible.
:::

## Preliminaries

The code used is shown above each plot and can be copied. We first load
the packages we will use.

When package 'ggpp' is loaded and attached, package 'ggplot2' is also
attached. The only function from 'ggplot2' that is redefined by 'ggpp'
is `annotate()`, which remains backwards compatible with 'ggplot2'.

```{r}
library(ggpp)
```

## Inset plots within plots

Inset plots are most frequently annotations. They can be used to zoom in
to a portion of the data, to show a less important property of
observations, or more commonly for maps, to show a broader region to
provide recognizable context.

::: callout-note
A future separate gallery is planned for map insets. They can be added similarly
as those shown here, but projections and coordinate systems add complexity.
:::

### Using `annotate()`

The first set of examples show how to use `annotate()` together with the
geoms from package 'ggpp' to add inset plots to single-panel ggplots.
The examples show different uses of insets. All examples use data sets
available in R or the packages loaded. In this examples we use
`annotate()` together with `geom_plot_npc()` or `geom_plot()`.

#### Zoom in

`geom_plot_npc()` obeys native plot coordinates instead of data
coordinates. NPC coordinates are relative to the plotting area and take
values in \[0..1\]. They make consistent positioning of annotations
easy. Only for this first example we show the use of both
`geom_plot_npc()` and `geom_plot()`. In both geoms the default for
`hjust` and `vjust` is `"inward"`.

```{r}
# zoom in region
zoom.region <- list(x = c(4.9, 6.2), y = c(13, 21))

# main plot
p.base <- ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point()

# inset zoomed-in plot
p.inset <- p.base + 
  coord_cartesian(xlim = zoom.region$x, ylim = zoom.region$y) +
  labs(x = NULL, y = NULL) +
  theme_bw(8) +
  scale_colour_discrete(guide = "none")

# whole plot using annotate() and NPC coordinates
p.combined <- p.base + 
  annotate(geom = "plot_npc", npcx = 0.95, npcy = 0.95, label = p.inset)

# highlight zoomed-in region
p.combined + 
  annotate(geom = "rect", 
           xmin = zoom.region$x[1], xmax = zoom.region$x[2], 
           ymin = zoom.region$y[1], ymax = zoom.region$y[2],
           linetype = "dotted", fill = NA, colour = "black")
```

`geom_plot()` obeys the usual `x` and `y` aesthetics expressed in data
coordinates. As shown here, while NPC coordinates are refered to the
plotting area set by other plot layers, using data cordinates, just as
for text labels, results in its expansion as needed to include the
inset's *x* and *x* coordinates.

```{r}
# whole plot using annotate() and data coordinates
p.d.combined <- p.base + 
  annotate(geom = "plot", x = 7, y = 50, label = p.inset)

# highlight zoomed-in region
p.d.combined + 
  annotate(geom = "rect", 
           xmin = zoom.region$x[1], xmax = zoom.region$x[2], 
           ymin = zoom.region$y[1], ymax = zoom.region$y[2],
           linetype = "dotted", fill = NA, colour = "black")

```

#### Zoom out

```{r}
# zoom in region
zoom.region <- list(x = c(4.9, 6.2), y = c(13, 21))

# a plot with zoom area highlighted
p <- ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point() + 
  annotate(geom = "rect", 
           xmin = zoom.region$x[1], xmax = zoom.region$x[2], 
           ymin = zoom.region$y[1], ymax = zoom.region$y[2],
           linetype = "dotted", fill = NA, colour = "black")

# inset whole-plot 
p.inset <- p  +
  labs(x = NULL, y = NULL) +
  theme_bw(8) +
  scale_colour_discrete(guide = "none")

# main zoomed-in plot
p.base <- p + 
  coord_cartesian(xlim = zoom.region$x, ylim = zoom.region$y) +
  expand_limits(y = 24) 

# whole plot using annotate()
p.base + 
  annotate(geom = "plot_npc", npcx = 0.98, npcy = 0.98, label = p.inset) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.20)))

```

#### Summary plot

```{r}
# column plot of summaries as inset
p.inset <- 
  ggplot(mpg, aes(factor(cyl), hwy, fill = factor(cyl))) +
  stat_summary(geom = "col", fun = mean, width = 2/3) +
  labs(x = "Number of cylinders", y = NULL, title = "Means") +
  scale_fill_discrete(guide = "none")

# scatter plot as main plot
p.base <- 
  ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point() +
  labs(x = "Engine displacement (l)", y = "Fuel use efficiency (MPG)",
       colour = "Engine cylinders\n(number)") +
  theme_bw()

# plot with inset plot using annotate()
p.base +
  annotate("plot_npc", npcx = 0.97, npcy = 0.98, label = p.inset + theme_bw(8))

```

#### Other data or variables

```{r}
# box plot as inset
p.inset <- 
  ggplot(mpg, aes(factor(cyl), hwy, fill = factor(cyl))) +
  stat_boxplot() +
  labs(x = "Engine cylinders (number)", y = "") +
  scale_fill_discrete(guide = "none")

# scatter plot as main plot
p.base <- 
  ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point() +
  labs(x = "Engine displacement (l)", y = "Fuel use efficiency (MPG)",
       colour = "Engine\ncylinders\n(number)") +
  theme_bw()

# plot with inset plot using annotate()
p.base +
  annotate("plot_npc", 
           npcx = 0.97, npcy = 0.98, 
           label = p.inset + theme_bw(9), 
           vp.width = 0.6, vp.height = 0.4) # relative size of inset
```

### Using `geom_plot()` and `geom_plot_npc()`

In the examples above we used `annotate()` as it is the easiest
approach. However, `annotate()` does not support adding different
annotations to different panels in a faceted plot. Very small plots can
sometimes function as data labels, specially in maps. In both cases we
need to use the geometries directly and pass a data frame as argument to
their `data` formal parameter.

When building `data` object containing the inset plots we need to add a
list as a column in a `data.frame`. This is possible as long as the
lengths match, but the `list` object needs to be protected with `I()`,
R's identity function.

```{r}
# faceted scatter plot as base plot
p.base <- 
  ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point() +
  labs(x = "Engine displacement (l)", y = "Fuel use efficiency (MPG)",
       colour = "Engine\ncylinders\n(number)") +
  facet_wrap(facets = vars(cyl), labeller = label_both) +
  theme_bw()

# ad hoc function for constructing box plots
make_inset <- 
  function(target.cyl) {
    ggplot(subset(mpg, cyl == target.cyl),
           aes(factor(cyl), hwy)) +
      stat_boxplot() +
      labs(x = "Engine cylinders\n(number)", y = "") +
      scale_fill_discrete(guide = "none") + 
      theme_bw(7)
  }

# list of box plots as insets
p.inset.list <- lapply(unique(mpg$cyl), make_inset)

# data frame to be used as data for plotting insets
inset.df <- data.frame(cyl = unique(mpg$cyl),
                       plots =  I(p.inset.list),
                       npcx = c(0.97, 0.97, 0.0, 0.97),
                       npcy = 0.5)

# plot with inset plots using geom_plot_npc()
p.base +
  geom_plot_npc(data = inset.df, 
                vp.height = 0.8,
                mapping = aes(npcx = npcx, npcy = npcy, label = plots))
```

When using tibbles instead of data frames, the call to `I()` can be
omitted.

```{r}
# tibble to be used as data for plotting insets
inset.tb <- 
  tibble::tibble(cyl = unique(mpg$cyl),
                 plots =  p.inset.list,
                 npcx = c(0.97, 0.97, 0.0, 0.97),
                 npcy = 0.5)

# plot with inset plots using geom_plot_npc()
p.base +
  geom_plot_npc(data = inset.tb, 
                vp.height = 0.8,
                mapping = aes(npcx = npcx, npcy = npcy, label = plots))
```

## Inset tables within plots

Tables are less frequently used as insets than plots. As annotations,
they can be useful in two situations: 1) reporting results for the
statistical analysis of the plotted data, and 2) adding numerical
summaries about the data. Very small tables can be also useful as data
labels to highlight specific observations in a plot or data for
locations in a map.

### Summaries

### ANOVA tables

### As data labels

## Graphical objects as insets

### Bitmaps

### Vector graphics
