---
title: "Digital light sensors"
subtitle: "Miniature mutichannel integrated circuits"
date: 2025-08-19
date-modified: 2026-02-01
author: "Pedro J. Aphalo"
keywords: [light sensors, multichannel, digital]
categories: [measurement and control]
editor: 
  markdown: 
    wrap: 72
code-fold: true
format:
  html: 
    resource-path: 
      - ./sensor-images
    code-link: true
    code-tools: true
abstract: |
  Brief description of modern multichannel digital light sensors as components suitable for the construction of complete weather-proof sensors.
draft: false
---

```{r, message=FALSE}
library(ggspectra)
library(photobiologyWavebands)
library(photobiologySensors)

energy_as_default()
theme_set(theme_bw())
```

```{r, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4,
               cache = FALSE, message = FALSE)
options(warnPartialMatchArgs = FALSE)
```

## Modern digital light sensors

In the past, and still currently, most light sensors used for quantification of photosynthetically active radiation, for spectific bands within PAR and for FR are constructed based on discrete silicon photodiodes and optical filters, delivering a low voltage analogue signal. 
In recent years integrated circuits (ICs) based on CMOS technology including photodiodes, analogue amplification, analogue to digital conversion (ADC) and supporting standardised serial digital communication protocols have been developed for different purposes. Some of them are good enough to replace with advantage discrete photodiodes.

One of the key advances is the depostion of interferece filters directly on the IC chip, selectively on individual photodiodes. The incorporation of signal amplification makes it possible to rely on smaller photodiodes and by implementing gain adjustments increase the dynamic range. Furthermore, the implementation of an ADC in the same chip and relying on digital communication avoids noise pickup and makes possible the use of much simpler and cheaper circuits to acquire and log the data. Miniaturization also reduces power requirememts.

A key supplier this type of digital sensors is ams OSRAM that offers many different variations aiming at different applications. Vishay has also developed some sensors with a limited number of channels. All these sensors are extremely small and encapsulated as surface mounted devices (SMD) in very small packages (e.g., $2 \times 3 \times 1$ mm).

When bought as components in quantity they are very cheap (3 to 15 €). Given their size, for prototyping and custom assembly they are easier to use if bought as "break out boards" (10 to 25 €), i.e., soldered on a small printed circuit board (PCB) to easy the task of connecting them. They cannot be used on their own, they need to be connected to a micro-controller or single-board-computer to acquire the data and control their settings. 

The Swiss company YoctoPuce has developed USB modules containing a microcontroller, memory, an isolated power supply and a USB interface. These modules have a built in data logger and a webserver-based user interface as well as a programming API. Through a hub they can be accessed remotely through the Internet or LAN. These modules cost more than the bare sensor ICs (40 to 100 €) but provide a readily usable solution. Their Yocto-Spectral USB module based on ams OSRAM's AS743 digital spectral sensor when used for measuring irradiance require the addition of a diffuser. This module has built-in LEDs for measuring reflectance. 

Aranet has just released a weather proof sensor with a cosine diffuser and a LoRa interface that connects through a base station, likely based on a digital spectral sensor. This sensor is sold for 1000 € + VAT, and although the easiest to use being a ready made field sensor, it lacks flexibility in the precomputed data summaries or access to raw data from individual channels.

For all these sensors, the wavelength and channel sensitivity guaranteed in specifications have rather loose tolerances. In most cases electronic components are closer to the typical specifications than the tolerance limits.

Examples of promising digital sensors that could be useful in agriculture and other fields are given below.

::: callout-note
# UV Sensor VML6075 from Vishay

The VML6075 from Vishay is a digital UV sensor with two channels, named UVB and UVA in the documentation but closer to UVA1 and UVA2 in reality. Vishay provides an algorithm to compute UVI. This sensor is still available from distributors but its manufacture has been discontinued by Vishay.

```{r}
autoplot(sensors.mspct$Vishay_VEML6075, 
         w.band = UV_bands("CIE"), range = c(280, 450))
```
:::

::: callout-note
# UV Sensor AS7331 from ams OSRAM

The AS7331 from ams OSRAM is a digital UV sensor with three channels, named UVC, UVB and UVA in the documentation. The band separation is rather good for computing UVI in sunlight and the usual artificial UV
radiation sources.

```{r}
autoplot(sensors.mspct$ams_AS7331, 
         w.band = UV_bands("CIE"), range = c(200, 450))
```

Sensitivity of Si photodiodes to UV radiation decreases with decreasing wavelength, but the design compensates for this using a photodiodes of larger area to compensate.

:::

::: callout-note
# Spectral VIS and IR sensor AS7343 from ams OSRAM

The AS7343 from ams OSRAM is a digital UV sensor with 13 channels. This sensor seems like the most useful for measuring light from the perspective of plants as it seems possible to estimate PAR, ePAR and FR.

```{r}
autoplot(sensors.mspct$ams_AS7343, 
         w.band = VIS_bands())
```

Sensitivity of Si photodiodes to UV radiation decreases with decreasing wavelength, but the design compensates for this using a photodiodes of larger area to compensate. Wavelength and channel sensitivity guaranteed specifications are rather loose.

:::

::: callout-note
# Spectral VIS and IR sensor AS7341 from ams OSRAM

The AS7341 from ams OSRAM is a digital spectral sensor with 10 channels. The band separation is rather good for computing UVI in sunlight and the usual artificial UV
radiation sources.

```{r}
autoplot(sensors.mspct$ams_AS7341, 
         w.band = VIS_bands())
```

Sensitivity of Si photodiodes to UV radiation decreases with decreasing wavelength, but the design compensates for this using a photodiodes of larger area to compensate. Wavelength and channel sensitivity guaranteed specifications are rather loose.

:::

::: callout-note
# Spectral VIS and IR sensor TSL2591 from ams OSRAM

The TSL2591 from ams OSRAM is a digital illuminance sensor with two channels, named ch0 and ch1 in the documentation. Illuminance can be obtained by subtracting the NIR sensed by ch1 from ch0.

```{r}
autoplot(sensors.mspct$ams_TSL2591, 
         w.band = VIS_bands())
```

Sensitivity of Si photodiodes to UV radiation decreases with decreasing wavelength, but the design compensates for this using a photodiodes of larger area to compensate. Wavelength and channel sensitivity guaranteed specifications are rather loose.

:::

## Conclusion

Some of these very cheap sensors could be adapted to use in the measurement or estimation of quantities used in both agriculture practice and in research. To realize this potential methods for the reliable retrieval of the quantities of interest consistently in time and space need to be developed.

I am currently exploring possible uses of the Yocto-Spectral USB module in agriculture production and plant science research.
