---
title: "Digital light sensors"
subtitle: "Miniature multichannel integrated circuits"
date: 2025-08-19
date-modified: 2026-06-03
author: "Pedro J. Aphalo"
keywords: [light sensors, multichannel, digital]
categories: [measurement and control]
editor: 
  markdown: 
    wrap: 72
code-fold: true
format:
  html: 
    resource-path: 
      - ./sensor-images
    code-link: true
    code-tools: true
abstract: |
  Brief description of a selection of modern multichannel digital light sensors available as electronic components suitable for the construction of complete, possibly weather-proof, sensors. This description emphasizes their optical properties.
draft: false
---

```{r, message=FALSE}
library(ggspectra)
library(photobiologyWavebands)
library(photobiologySensors)

energy_as_default()
theme_set(theme_bw())
```

```{r, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4,
               cache = FALSE, message = FALSE)
options(warnPartialMatchArgs = FALSE)
```

## Modern digital light sensors

In the past, and still currently, most light sensors used for quantification of photosynthetically active radiation, for specific bands within PAR and for FR are constructed based on discrete silicon photodiodes and optical filters, delivering a low voltage analogue signal.

In recent years integrated circuits (ICs) based on CMOS technology including photodiodes, analogue amplification, analogue to digital conversion (ADC) and supporting standardized serial digital communication protocols have been developed for different purposes. Some of them are good enough to replace with advantage discrete photodiodes. They can be connected directly to a microcontroller or a single-board-computer.

A recent advance in the design of solid-state integrated "light" sensors has been the deposition of interference filters directly on the IC chip, selectively on individual photodiodes. Together with on-chip analogue signal amplification it makes possible the use of small photodiodes. As the amplification gain can be adjusted, the dynamic range is enhanced. Furthermore, including an analogue to digital converter (ADC) in the same chip and relying on digital communication avoids noise pickup in the wiring. Miniaturization also reduces power requirements.

Integrated digital sensors, even if calibrations are frequently applied during manufacture, aim to be cheap and thus their specifications can have wider wavelength tolerances under test conditions. However, some of these new IC sensors have temperature compensation and a temperature sensor also integrated resulting in extremely low and consistent dark noise readings making high amplification gain possible, contributing to an enhanced dynamic range.

All these sensors are extremely small and encapsulated as surface mounted devices (SMD) in very small packages (e.g., $2 \times 3 \times 1$ mm) with a tiny window. As they are based on interference filters their angle of acceptance is narrow, in many cases only 20 degrees.

Analogue amplification gain and the integration time in the ADC are set using the same communication link as for transmitting data. Thus, the instructions controlling the settings are sent by the same microcontroller or microcomputer board used to collect the data. Miniaturization and the use of low voltages makes these sensors very energy efficient, but the length or wiring between the sensor IC and the microcontroller is limited to a few centimetres.

A key supplier of digital light sensor ICs is [ams OSRAM](https://ams-osram.com/products/sensor-solutions/ambient-light-color-spectral-proximity-sensors/) that offers many different multichannel sensors aiming at different applications. The main differences are in the number of channels and their wavelength sensitivity.  [Vishay](https://www.vishay.com/en/optoelectronics/) has also developed similar sensors with a limited number of channels. 
When bought as components in quantity they are very cheap (3 to 15 €). Given their size, for prototyping and custom assembly they are easier to use if bought as "break out boards" (10 to 25 €), i.e., ready soldered on a small printed circuit board (PCB) to easy the task of connecting them. These sensor, even when soldered on break out boards, cannot be used on their own. They need to be connected to a micro-controller or single-board-computer to acquire the data and control their settings. 

The Swiss company [YoctoPuce](https://www.yoctopuce.com/) has developed USB modules containing a microcontroller, memory, an isolated power supply and a USB interface. These modules have a built in data logger and a webserver-based user interface as well as a programming API. Through a hub they can be accessed remotely through the Internet or LAN. These modules cost more than the bare sensor ICs (40 to 100 €) but provide a readily usable solution. Their [Yocto-Spectral USB module](https://www.yoctopuce.com/EN/products/usb-environmental-sensors/yocto-spectral) based on ams OSRAM's AS743 digital spectral sensor when used for measuring irradiance require the addition of a diffuser. This module has built-in LEDs for measuring reflectance. 

Aranet has just released a weather proof sensor with a cosine diffuser and a LoRa interface that connects through a base station, likely based on a digital spectral sensor. This sensor is sold for 1000 € + VAT, and although the easiest to use being a ready made field sensor, it lacks flexibility in the precomputed data summaries or access to raw data from individual channels.

For all these sensors, the wavelength and channel sensitivity guaranteed in specifications have rather loose tolerances. However, in most cases electronic components are closer to the typical specifications than the tolerance limits.

Examples of digital sensors that I think can be useful in agriculture and in related research fields such as ecology and physiology of plants are given below.

::: callout-note
# UV Sensor VML6075 from Vishay

The VML6075 from Vishay is a digital UV sensor with two channels, named UVB and UVA in the documentation but closer to UVA1 and UVA2 in reality. Vishay provides an algorithm to compute UVI. This sensor is still available from distributors but its manufacture has been discontinued by Vishay.

```{r}
autoplot(sensors.mspct$Vishay_VEML6075, 
         w.band = UV_bands("CIE"), range = c(280, 450))
```
:::

::: callout-note
# UV Sensor AS7331 from ams OSRAM

The AS7331 from ams OSRAM is a digital UV-radiation sensor with three channels, named UVC, UVB and UVA in the documentation. The band separation is rather good for computing UVI in sunlight and the usual artificial UV radiation sources.

```{r}
autoplot(sensors.mspct$ams_AS7331, 
         w.band = UV_bands("CIE"), range = c(200, 450))
```

Sensitivity of Si photodiodes to UV radiation decreases with decreasing wavelength, but the design compensates for this using a photodiodes with larger area in this sensor than in those for VIS radiation.

:::

::: callout-note
# Spectral VIS and IR sensor AS7343 from ams OSRAM

The AS7343 from ams OSRAM is a digital VIS and NIR sensor with 13 channels. This sensor seems like the most useful for measuring light from the perspective of plants as it seems possible to estimate PAR, ePAR and FR.

```{r}
autoplot(sensors.mspct$ams_AS7343, 
         w.band = VIS_bands())
```

:::

::: callout-note
# Spectral VIS and IR sensor AS7341 from ams OSRAM

The AS7341 from ams OSRAM is a digital spectral VIS and IR sensor with 10 channels. The channels have wavebands of similar width suggesting that it could work well as a spectrometer for visible light. From the perspective of measuring light as relevant to plants, it lacks a far-red channel.

```{r}
autoplot(sensors.mspct$ams_AS7341, 
         w.band = VIS_bands())
```

:::

::: callout-note
# Spectral VIS and IR sensor TSL2591 from ams OSRAM

The TSL2591 from ams OSRAM is a digital illuminance sensor with two channels, named ch0 and ch1 in the documentation. Illuminance can be obtained by subtracting the NIR sensed by ch1 from ch0.

```{r}
autoplot(sensors.mspct$ams_TSL2591, 
         w.band = VIS_bands())
```

:::

::: callout-note
# Spectral VIS sensor AS7263 from ams OSRAM

The AS7262 from ams OSRAM is a digital VIS sensor with 6 channels. This sensor has regularly spaced wavelength bands with a FWHM of 20 nm.

```{r}
autoplot(sensors.mspct$ams_AS7263, 
         w.band = VIS_bands())
```

:::

::: callout-note
# Spectral IR sensor AS7263 from ams OSRAM

The AS7263 from ams OSRAM is a digital red + NIR sensor with 6 channels. This sensor has been used in instruments for estimation of NDVI.

```{r}
autoplot(sensors.mspct$ams_AS7263, 
         w.band = VIS_bands())
```

:::

## Conclusion

Some of these very cheap sensors could be adapted to use in the measurement or estimation of quantities used in both agriculture practice and in research. To realize this potential methods for the reliable retrieval of the quantities of interest consistently in time and space need to be developed.

I am currently exploring possible uses of the Yocto-Spectral USB module in agriculture production and plant science research.
