---
title: "Walz LSA-2050"
subtitle: "First impressions"
author: "Pedro J. Aphalo"
date: 2026-05-07
date-modified: 2026-05-07
keywords: [Walz LSA-2050]
categories: [measurement and control]
editor: 
  markdown: 
    wrap: 72
code-fold: true
format:
  html: 
    code-link: true
    code-tools: true
abstract: |
  I desccribe here my first impressions about the Walz LSA-2050 handheld leaf analyzer. These impressions are based on long-term experience with the use of the SPAD-502 chlorophyll meter from Konica-Minolta, the Dualex 3.1 and Dualex 4 dual-excitation instruments from Force-A and recent use of the MPM-100 multiple wavelength pigment meter from Opti-Sciences . I also show how to import the data logged by these instruments into R for further analysis.
draft: true
---

## The LSA-2050 and its altrenatives

In this section I describe the instruments from newest to oldest.

The first impression of the *Walz LSA-2050* is that of a refined design
and very well tested instrument, making it reliable, ergonomically easy
to use and with very well thought out and clear firmware user interface.
In some respects it benefits from Walz's long experience in chlorophyll
fluorescence measurements, as well as from the approaches used in the
Dualex instruments. The *LSA-2050* does the measurements very fast. With
the help from its good ergonomics, I was able to measure 15 leaves, each
from a different plant, across four growth-chamber shelves in 4 min.

The *MPM-100* is not really a clone of the Dualex in its measurement
approach but attempts to be a replacement for it. At least the version I
have used, has a display with very small text, even if in colour, and
some significant rough edges in the firmware user interface. The most
important being that when a measurement fails, the screen displays the
results from the last good measurement, which is prone to cause mistakes
in the manual recording of readings. Another quirk is that when the
estimation of one parameter fails the Dualex reports those that could be
measured, instead, the MPM-100 returns no results, which is problematic.
The MPM-100 is also slower than the Dualex as it averages several
consecutive measurements.

The *Dualex 4* is also a well designed instrument, a refinement of the
earlier *Dualex 3*. It has a couple weak points, though: the hinge is
rather weak and with long-term can develop play or even break. In the
first case some misalignment of the two halves of the measurement head
can occur. The Dualex has more options for encoding treatments than the
Walz, making it more cumbersome, but more flexible in the tagging of
logged data.

The *SPAD-502* is no longer available in the simple version I have used.
The *SPAD-502Plus* with data logging capabilities is also a well
designed and easy to use instrument. I have not used any of the cheaper
clones.

I should mention here also the *SpectraVue Leaf Spectrometer* from CID
Bio-Science. This is a **badly designed** instrument that returns **bad
data**. It should not be used in scientific research. First it suffers
from extreme problems of dark noise and stray light, as can be seen in
the images in the company's own advertising. More importantly, because
of the light source used and the configuration of the instrument
entrance optics the sum of reflectance, transmittance and absorptance
reported is very far from the theoretical value of one, but much larger,
i.e., the values reported as reflectance, transmittance and absorptance,
are not these physical quantities! (The company has been aware of these
design problems for several years, as they recognized when I reported
them several years ago in a exchange of several e-mails.)

## Chlorophyll

*In vivo* measurement of chlorophyll based on estimating leaf
transmittance return concentrations expressed per unit leaf area. Given
that many variables in the growing environment affect specific leaf area
(SLA) differences in chlorophyll concentration between treatments depend
on the base of expression. In practice, the direction of a response can
be opposite when concentration is expressed per unit area and unit dry-
or fresh-biomass.

Chlorophyll is estimated based on the transmittance of the leaf at
specific wavelengths where the pigment of interest in the main absorber
of light. A leaf differs from a solution in that the chlorophyll
distribution is not homogeneous. Chlorophyll is in the chloroplasts, and
the location of chloroplasts within mesophyll cells changes in response
to illumination. The changes in transmittance can even be used to study
chloroplast accumulation movements [@justyna]. Thus, the relationship
between chlorophyll concentration and light transmittance varies
depending on illumination, specially blue light sensed by phototropins
[@xxxx]. This has been described as a source of errors in chlorophyll
measurements with the SPAD-502 instrument [@xxxx], but as this is an
optical phenomenon in the plant it can be expected to affect of the
instruments described here.

The concentration of chlorophyll together with the lengthened path of
photons inside a leaf due to scattering, makes of leaves very effective
absorbers such that transmittance at the wavelength of maximum
absorption by chlorophyll is extremely low, making the measurement of
high concentrations of chlorophyll with instruments like the SPAD-502
error-prone. The Dualex 4 introduced the approach of using a longer
wavelength, were transmittance is higher but still mainly dependent on
chlorophyll and little affected by other pigments and metabolites. There
exits two versions of the MPM-100 instrument, the MPM-100 using an
approach and wavelengths similar to the Dualex and the MPM-100/S that
estimates chlorophyll using an approach similar to the SPAD-502.

The light scattering within leaves decreases the apparent transmittance
even in the absence of absorbing pigments. This effect is estimated by
measuring the apparent transmittance of far-red or near-infrared light
by the leaf, as at these wavelengths absorption by pigments is minimal.

The readings are expected not to depend on which side of the leaf is
illuminated. In fact, the Dualex 4 and the LSA-2050, illuminate opposite
epidermis and use the same sensor for fluorescence and transmittance.
Like the Dualex, the MPM-100 measures transmittance with illumination on
the "measured" leaf side, but using different sensors for transmittance
and fluorescence.

## Epidermal pigments

Fluorescence is dependent on the number of absorbed photons, this also
applies to chlorophyll. Thus, a measurement of fluorescence behind the
epidermis, from the chlorophyll in the mesophyll, can be used to
estimate the transmittance of the epidermis if it is possible to
estimate the fluorescence yield that can be expected in the absence of
the epidermis. One approximation is to use a wavelength that is known
not to be absorbed by the epidermis to  a reference value of
fluorescence to compare against fluorescence at other wavelengths where
the epidermis can absorb.

This the principle of the Dualex, similar to the earlier approach used
by the UVA-PAM, a modified Walz Mini-PAM. The Dualex, however, measures
the fluorescence on the opposite side of the leaf from the excitation,
while the UVA-PAM and the Walz LSA-2050 measure the fluorescence from
the side of the leaf where the excitation light impings. While the
Dualex can be used only on thin leaves, the LSA-2050 can be used on
thicker objects like fruits (by removing the lower half of the head,
which contains the light sources for estimation of chlorophyll based on
transmittance across the leaf)

The MPM-100 senses the chlorophyll fluorescence on the same side of the
leaf as the excitation impings. However the head has a maximum opening
that limits its use to thin objects like leaves.

## Excitation wavelengths

All the instruments described use LEDs as sources for excitation of
fluorescence and for measurement of transmittance. The
instruments differ in the excitation wavelengths used for assessing
epidermal transmittance and in their number. The wavelengths used
for assessing chlorophyll concentration based on whole-leaf transmittance
also vary.

| Instrument  | UV-B     | UV-A1    | Blue     | Green    | Red      | Chl                 |
|:-----------|:-----------|:-----------|:-----------|:-----------|:-----------|:-----------|
| LSA-2050    | 310 (15) | 365 (12) | 450 (14) | 530 (27) | 630 (24) | 715 (25) + 770 (30) |
| MPM-100     |          | 375      |          | 525      | 660      | 720 + 850           |
| MPM-100/S^*^ |         | 375      |          | 525      | 660      | 650 + 940           |
| Dualex 4    |          | 375      |          | 520      | 630      | 720 + 810?          |
| Dualex 3 FLAV |        | 375      |          |          | 630      |                     |
| Dualex 3 ANTH^*^ |     |          |          | 520      | 630      |                     |
| Dualex 3 CA | 310      |          |          |          | 630      |                     |
| UVA-PAM^*^     |       | 375      | 470      |          |          |                     |
| UVA-PAM/red^*^ |       | 375      |          |          | 630?     |                     |
| Xe-PAM^*^ | 314 (18)   | 366 (32) | 475 (140) |         |          |                     |
| SPAD-502    |          |          |          |          |          | 650 + 940           |

: Excitation wavelengths used by the different instruments, full width half maximum (FWHM) given in parentheses when available. Instruments names followed by ^*^ indicate that the author has no direct experience in their use.

## Cost

The *SPAD-502Plus* seems to cost about 2500-3000 €, with various clones
available in the range 600-1500 €. The *Dualex 4*, now from Pessl after the bankruptcy of Force-A,
costs close to 4500 €, the *MPM-100* costs close to 3000 € and the *Walz
LSA-2050* costs close to 6000 €. (These prices are very approximate as
they vary depending on supplier, import taxes, etc., as well as with
accessories ordered and in some cases configurations).

## Comparisons in the literature

## Data logging and further processing

The LSA-2050 logs measurements to its internal memory and these data can be downloaded as an Excel workbook using a Windows program provided for free with the instrument and also available for download in its current version. The worksheets contain both hidden and visible values. Many of the values are computed by formulas in the worksheet, something that makes import more complex. The worksheets as returned by the LSA software do  seem to contain bare formulas, without cached computed values.

The computed values are necessary for import into foreign software such as R. Thus, the calculated values need to be added to the worksheets. The workbook is compatible with LibreOffice, but in contrast to Excel that automatically calculates the values on file load, in LibreOffice a hard recalculation needs to be manually triggered to have the computed values displayed instead of zeros. After resaving the file it can be imported into R. At the moment, Linux and Mac OS X users are out-of-luck as no software for data download is yet available.

There are multiple R packages capable of reading Excel worksheets from workbooks. The first package I tried was the popular 'readxl' and it worked on my first attempt. However, the data are in two separate worksheets within a workbook.

::: callout-attention
When computed values for formulas are not present, the cells are displayed filled with zeros. If precomputed values are lacking reading the file into R also results is the import of zeros instead of the computed values.
:::

The data in the "SAT chart" worksheet are time series with time points along rows, which is not how plotting functions of R expect the data to be organized. Because of this, even if the Excel worksheets can be easily read into R, additional processing is needed. In addition, in many cases only the computed values are needed for statitical analysis and plotting in R, making subsetting necessary after import.

I have added function `read_walz_lsa_xlsx()` to package 'photobiologyInOut'. This function reads the worksheets, checks that the computed values are available, and if not replaces the filler zeros with `NA` markers with a warning. The data from both worksheets are combined into a single data frame. The SAT time series are stored in one data frame per sample, nested inside the main data frame containing all raw and computed data for each sample in one row. Different subsets of the data in the workbook can also be imported when not all data are needed.

### Example

```{r}
library(dplyr)
library(photobiologyInOut)
library(ggplot2)
library(ggpmisc)
library(patchwork)
```

```{r}
faba_light_before.df <-
  read_walz_lsa_xlsx("lza-dualex-mpm-data/faba-light-before.xlsx")
faba_light_before.df$condition <- "light.240min"
```

```{r}
faba_20min_dark.df <-
  read_walz_lsa_xlsx("lza-dualex-mpm-data/faba-20min-dark.xlsx")
faba_20min_dark.df$condition <- "dark.20min"
```

```{r}
faba_35min_dark.df <-
  read_walz_lsa_xlsx("lza-dualex-mpm-data/faba-35min-dark.xlsx")
faba_35min_dark.df$condition <- "dark.35min"
```
  
```{r}
faba_light_after.df <-
  read_walz_lsa_xlsx("lza-dualex-mpm-data/faba-light-after.xlsx")
faba_light_after.df$condition <- "light.10min"
```

```{r}
faba_191_LSA.df <-
  bind_rows(faba_light_before.df, 
            faba_20min_dark.df,
            faba_35min_dark.df,
            faba_light_after.df) |>
  mutate(condition = factor(condition),
         condition = reorder(condition, Time))
```

```{r}
ggplot(faba_191_LSA.df, aes(condition, nmol.cm2)) +
  geom_boxplot() +
  stat_multcomp(label.type = "letters", size = 2.7) +
  expand_limits(y = 0)
```

```{r}
ggplot(faba_191_LSA.df, aes(condition, AFLAV)) +
  geom_boxplot() +
  stat_multcomp(label.type = "letters", size = 2.7) +
  expand_limits(y = 0)
```

```{r}
ggplot(faba_191_LSA.df, aes(condition, AANTH)) +
  geom_boxplot() +
  stat_multcomp(label.type = "letters", size = 2.7) +
  expand_limits(y = 0)
```

```{r}
ggplot(faba_191_LSA.df, aes(condition, FV.FM)) +
  geom_boxplot() +
  stat_multcomp(label.type = "letters", size = 2.7)
```


```{r}
ggplot(faba_191_LSA.df, aes(condition, Fo)) +
  geom_boxplot() +
  stat_multcomp(label.type = "letters", size = 2.7)
```

```{r}
ggplot(faba_191_LSA.df, aes(condition, Fm)) +
  geom_boxplot() +
  stat_multcomp(label.type = "letters", size = 2.7)
```

```{r}
ggplot(faba_191_LSA.df, aes(condition, Fm - Fo)) +
  geom_boxplot() +
  stat_multcomp(label.type = "letters", size = 2.7)
```

```{r}
ggplot(faba_191_LSA.df, aes(nmol.cm2, AANTH)) +
  geom_point() +
  stat_correlation(use_label("r", "r.confint", "P"),
                   label.x = "right") +
  stat_ma_line() +
  stat_ma_eq(use_label("eq"), label.y = "bottom") +
  expand_limits(y = 0)
```

```{r}
n <- 6 # "plant"

(ggplot(faba_light_before.df$SAT.F.ls[[n]], aes(t, SAT.F)) +
  geom_point() + 
  geom_line() +
  expand_limits(y = c(0, 1200))) +
(ggplot(faba_20min_dark.df$SAT.F.ls[[n]], aes(t, SAT.F)) +
  geom_point() + 
  geom_line() +
  expand_limits(y = c(0, 1200))) +
(ggplot(faba_35min_dark.df$SAT.F.ls[[n]], aes(t, SAT.F)) +
  geom_point() + 
  geom_line() +
  expand_limits(y = c(0, 1200))) +
(ggplot(faba_light_after.df$SAT.F.ls[[n]], aes(t, SAT.F)) +
  geom_point() + 
  geom_line() +
  expand_limits(y = c(0, 1200)))
```
