---
title: "Geometries geom_text_s() and geom_label_s()"
subtitle: "ggpp: Grammar Extensions to 'ggplot2'"
author: "Pedro J. Aphalo"
date: 2022-12-17
date-modified: 2026-04-23
aliases:
  - ../posts/Enhancing-geom-text/index.html
bibliography: ../datviz-ggplot-and-friends.bib
format:
  html:
    code-fold: true
    code-tools: true
categories: [R, plotting, ggpp, ggplot2]
keywords: [ggpp pkg, ggplot2 pkg, data visualization, dataviz]
abstract: |
  Geoms geom_text_s() and geom_label_s() from 'ggpp' are enhanced versions of geom_text() and geom_label() from 'ggplot2'. In this web page I discuss what the enhancements are and what is the rationale behind their design. I demonstrate with examples how the enhanced features in geom_text_s() and geom_label_s() can be used to add more effective and visually attractive data labels to ggplots.
---

::: callout-warning
Versions of package ['ggpp'](https://docs.r4photobiology.info/ggpp) released before 0.5.0 included a preliminary attempt at the design of these geometries based on a different logic and interface. At the time of writing, version 0.5.2 is available through CRAN. To reproduce the examples at the end of this page please use 'ggpp' (>= 0.5.0) and preferably 'ggpp' (>= 0.5.2).
:::

## Selective mapping of aesthetics

I have defined in package ['ggpp'](https://docs.r4photobiology.info/ggpp/) enhanced versions of `geom_text()` and `geom_label()` under the names of `geom_text_s()` and `geom_label_s()` . The `s` is for segment, at least it was when I thought of these names. The versions described are included in 'ggpp' (>= 0.5.0).

Data labels are text semantically connected to individual data observations or individual data summaries depicted on the plot canvas [@Koponen2019]. The idea of better supporting the use of data labels in plots created with R package ['ggplot2'](https://ggplot2.tidyverse.org/) is not new. I have been for a long time a happy user of the repulsive geometries `geom_text_repel()` and `geom_label_repel()` from package ['ggrepel'](https://ggrepel.slowkow.com/). I have even contributed some code to 'ggrepel'. However, some time back when designing some new position functions, the idea of developing non-repulsive geometries suitable for displaced data labels started growing on me. From a graphical design perspective, I also found at times `geom_text()` and `geom_label()` to have limitations.

Before version 4.0.0 of 'ggplot2', in `geom_text()` transparency set with `alpha` was applied to the text while in `geom_label()` it affected only `fill`. In `geom_label()` the `colour` aesthetics controlled at the same time the colour of the text and of the line bordering the label; `label.size` controlled the size of the border line, but as a parameter rather than an aesthetic. This meant that it was not possible to use the border width, its absence or its colour to highlight selected data labels. Lack of support for the linetype aesthetic of the border, made it difficult to highlight individual data labels in black and white printing. _Thus changed significantly in 'ggplot2' (>= 4.0.0)_. _However, use cases remain where the additional flexibility brought by the different approach used in 'ggpp' by `geom_text_s()` and `geom_label_s()` are either necessary, e.g., connecting segments and arrows, or at least more convenient, e.g., targeting of aesthetic mappings to specific elements of the graphical objects (grobs) created by the geoms._

When designing `geom_text_s()` and `geom_label_s()`, I aimed from the start to keep the use of any new geometries simple, and as consistent with the grammar of graphics as possible. The interfaces of `geom_text_s()` and `geom_label_s()` are stable. Thus, these geometries are very likely to remain backwards compatible in future versions of 'ggpp'. By default `geom_text()` and `geom_label()` behave (almost?) as their counterparts from 'ggplot2' (< 4.0.0). _They do not make use of new aesthetics as `geom_text_repel()` and `geom_label_repel()` do, instead, they rely on additional formal parameters to target the mapped aesthetics._

::: callout-note
# Rationale

The decision to use a targeting mechanism instead of duplicating existing aesthetics, is that multiple inconsistent mappings of variables to the same graphical property can make interpreting a plot difficult [@Koponen2019]. Targeting of aesthetics adds flexibility without weakening the Grammar of Graphics principle of allowing only a single mapping per aesthetic property within a plot, e.g., I see enforcing a single meaning for colour values within a plot as a fundamental design principle for data visualizations.   
:::

The enhancements are: 1) if used with one of the position functions from 'ggpp', which keep the original position, segments are drawn connecting the text or label to the observation. The justification, `vjust` and `hjust` and thus the anchoring point of the segment to the text or label are computed automatically based on the direction of displacement. 2) A new formal parameter `colour.target` makes it possible to control to which components of the graphical element (grob) to apply the _mapped colour aesthetic_ and to which elements to apply the _default colour_. 3) Similarly a new formal parameter `alpha.target` allows similar control of the application of the _mapped alpha aesthetic_ values and which ones obey the _default alpha_. Currently supported elements are `"text"`, `"segment"`, `"box.line"` and `"box.fill"` in any combination, and `"box"` and `"all"` as aliases to save typing. 4) Add support for `linewidth` and `linetype` aesthetics in `geom_label_s()` applying them to the box border line, independently of `segment.linewidth` which is a parameter. _The approach used in `geom_text()` and `geom_label()` from 'ggplot2' (>= 4.0.0) is different and not as flexible._

The use of segments or arrows to make clear to which data observation data labels are linked is frequent except in very sparse clouds of observations. Using by default *automatic* justification `"position"` based on the direction of displacement by nudging, dodging, etc., to select the anchor point ensures that short segments will not cause difficulties even with long text labels. It also means that small displacements away from observations are enough to avoid overlaps between label and the labelled observation. The `vjust` and `hjust` aesthetics except for the support of the additional value `"position"`, work as in 'ggplot2', accepting the same values as `geom_text()` and `geom_label()` and to the same effect.

In the case of these two geometries, the design stage was time-consuming and included attempts that in use resulted awkward. Implementation of the new features was straightforward. The examples below show some of the new possibilities for the graphical design of text label annotations in ggplots that `geom_text_s()` and `geom_label_s()` make possible.

_While the positions of labels when using `geom_text_repel()` and `geom_label_repel()` are determined during plot rendering and dependent on the theme and dimensions of the rendered plot as passed to graphic devices, in the case of `geom_text_s()` and `geom_label_s()`, similarly to with `geom_text()` and `geom_label()`, the positions are determined earlier and dependent on the mapped data and positions added to the `"gg"` object. Depending on the use case, repulsion during plot rendering can be an advantage or a hindrance._

## Examples

::: callout-note
# R code

Above each plot you will find the code used to create the figures, "folded" behind a triangle and the word Code. Clicking on the triangle or on "Code" makes the code visible/hidden. The code can be easily copied by clicking on the icon that appears when the cursor in on the upper right corner of the box.

The code used for the individual plots requires the code in the code chunk immediately under this call out to be run first, to attach the packages used, prepare the data and save a base plot with name `p` to which plot layers are added in later examples.
:::

```{r, message=FALSE}
#| code-fold: show
library(ggpp)
my.cars <- mtcars[c(TRUE, FALSE, FALSE, FALSE), ]
my.cars$name <- rownames(my.cars)
p <- 
  ggplot(my.cars, aes(wt, mpg, label = name)) +
  scale_colour_discrete(l = 40) # luminance, make colours darker
```

A simple example, using nudging to displace the labels.

```{r}
# Use nudging
p +
  geom_text_s(nudge_x = 0.12) +
  geom_point() +
  expand_limits(x = 6.2)
```

We add arrow points to the segments.

```{r}
p +
  geom_text_s(nudge_x = 0.15,
              arrow = arrow(length = grid::unit(1.5, "mm")),
              point.padding = 0.4) +
  geom_point() +
  expand_limits(x = 6.2)

```

Using colour with the default target, which, consistently with `geom_text()` is the text element of the composite grobs.

```{r}
p +
  geom_text_s(aes(colour = factor(cyl)),
              angle = 90,
              nudge_y = 1) +
  geom_point() +
  expand_limits(y = 27)
```

We can select a different graphic element as the target for the colour aesthetic, here we apply colour to the connecting segments.

```{r}
p +
  geom_text_s(aes(colour = factor(cyl)),
              colour.target = "segment",
              angle = 90,
              nudge_y = 1,
              show.legend = FALSE) +
  geom_point(aes(colour = factor(cyl))) +
  expand_limits(y = 27)
```

And here we apply colour to all the elements created by `geom_text_s()`.

```{r}
p +
  geom_text_s(aes(colour = factor(cyl)),
              colour.target = "all",
              angle = 90,
              nudge_y = 1) +
  geom_point() +
  expand_limits(y = 27)
```

In the case of `geom_label_s()` additional targets for colour are available. We start with the default, that is consistent with `geom_label()` applying the colour aesthetic to the border line and text.

```{r}
p +
  geom_label_s(aes(colour = factor(cyl)), nudge_x = 0.3) +
  geom_point() +
  expand_limits(x = 7)

```

We set the default colour to `"grey50"` to make segments and box borders less
prominent, achieving a calm design.

```{r}
p +
  geom_label_s(aes(colour = factor(cyl)),
               colour.target = "text",
               default.colour = "grey50",
               linewidth = 0.4,
               nudge_x = 0.3) +
  geom_point() +
  expand_limits(x = 7) +
  theme_bw()

```

We can use the opposite approach, keeping the text black for maximum legibility at a small size and apply colour to the remaining elements.

```{r}
p +
  geom_label_s(aes(colour = factor(cyl)), 
               size = 2.8,
               colour.target = c("box", "segment"),
               linewidth = 0.4,
               nudge_x = 0.3) +
  geom_point(aes(colour = factor(cyl))) +
  expand_limits(x = 7)

```

Aesthetic `linetype` can be used to unobtrusively highlight some labels in the previous plot.

```{r}
p +
  geom_label_s(aes(colour = factor(cyl), linetype = grepl("Merc", name)), 
               size = 2.8,
               colour.target = c("box", "segment"),
               linewidth = 0.4,
               nudge_x = 0.3,
               show.legend = FALSE) +
  geom_point(aes(colour = factor(cyl)), size = 3) +
  expand_limits(x = 7)

```

A visually simple and space-saving design, by combining fill and colour but suppressing the box border.

```{r}
p +
  geom_label_s(aes(colour = factor(cyl), fill = factor(cyl)),
               nudge_x = 0.2,
               size = 3.5,
               colour.target = "all",
               alpha.target = c("box", "segment"),
               alpha = 0.15,
               linewidth = 0,
               segment.linewidth = 1,
               label.padding = grid::unit(0.12, "lines")) +
  geom_point(aes(colour = factor(cyl)), size = 2) +
  expand_limits(x = 7) +
  theme_bw()
```

If we disable the plotting of segments, `geom_tetx_s()` behaves like `geom_text()` and `geom_label_s()` as `geom_label()` except for the flexibility in the mapping of `colour` and `alpha` aesthetics in `geom_label_s()` and the default justification `"position"`.

```{r}
p +
  geom_text_s(aes(colour = factor(cyl)),
              add.segments = FALSE,
              angle = 90,
              nudge_y = 0.3,
              show.legend = FALSE) +
  geom_point(aes(colour = factor(cyl))) +
  expand_limits(y = 27)
```

::: callout-tip
For additional information, please, see the [help page for geometries `geom_text_s()` and `geom_label_s()`](https://docs.r4photobiology.info/ggpp/reference/geom_text_s.html) and the [documentation of package 'ggpp'](https://docs.r4photobiology.info/ggpp), including a vignette.

Chapter 9 in @Aphalo2024 provides a short introduction to 'ggplot2', 'ggpp', 'ggpmisc' and other extensions.
:::
