---
title: "ggpmisc 0.5.0/0.5.1/0.5.2"
subtitle: "ggpmisc: Miscellaneous Extensions to 'ggplot2'"
author: "Pedro J. Aphalo"
date: 2022-08-05
date-modified: 2023-04-28
categories: [R Package Update]
keywords: [ggpmisc pkg]
---

::: callout-note

As part of moving the site to Quarto I intend to write a single post to give an
overview of changes in minor versions of the packages. In this case, all
versions in the 0.5.x series. I will update this same post in the case of minor
version updates, and start a new post when the first version in the 0.6.x series
is developed. I hope this will reduce the clutter and still provide a good
overview of progress. Differences between versions are listed in detail in the
[NEWS file](https://docs.r4photobiology.info/ggpmisc/news/).

:::

## Overview of changes

_Version 0.5.0_ brings several enhancements to the annotations based on model fits. The most visible change is the new convenience function `use_label()` that greatly simplifies the assembly of labels from components and their mapping to aesthetics. We exemplify its use and some of the other new features below in section Examples. Several of the model-fit based statistics return additional variables in the returned data adding flexibility. Function `stat_correlation()` now computes confidence intervals for correlation estimates. Add functions `keep_tidy()`, `keep_glance()` and `keep_augment()` as wrappers on methods `tidy()`, `glance()` and `augment()` from package 'broom'. These new functions make it possible to keep a trace of the origin of the _"broom-tidied"_ outputs.

_Version 0.5.1_ brings additional enhancements to the annotations based on model fits to improve traceability. New scales `scale_colour_logFC()`, `scale_color_logFC()` and `scale_fill_logFC()` and revised `scale_colour_outcome()` and `scale_fill_outcome()` add flexibility.

_Version 0.5.2_ fixes bugs and ensures compatibility with updates to 'ggplot2' and 'lubridate'. 

## Examples

In this section we demonstrate the use of some of the new features described above.

In the first plot we add an estimate of the correlation coeficient _R_, and the corresponding _t_-value and _P_-value.

```{r, message=FALSE}
library(ggpmisc)
```

```{r}
ggplot(subset(mpg, cyl != 5), aes(displ, hwy, colour = factor(cyl))) +
  geom_point() +
  stat_correlation(use_label(c("R", "t", "P")),
                   label.x = "right") +
  theme_bw()
```

The displacement volume of car engines is known without error and it can the thought as a possible explanation for the petrol use per distance (MPG or miles per gallon) for highway driving. We fit a linear regression, and annotate the plot with the fitted linear model equations, the _F_-values and _P_-values.

```{r}
ggplot(subset(mpg, cyl != 5), aes(displ, hwy, colour = factor(cyl))) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "F", "P")),
                   label.x = "right") +
  theme_bw()
```

In the case of the petrol use (MPG) in city traffic compared to highway travel the two variables can be expected to be subject to similar error variation and there is no directional cause-effect relationship between them. So, in this case OLS linear regression is not a suitable approach. We use major axis regression instead, and we add to the label, θ, the angle in degrees between the two lines that could have been fitted by linear regression using _x_ or _y_ as the explanatory variable.

```{r}
ggplot(subset(mpg, cyl != 5), aes(cty, hwy)) +
  geom_point(alpha = 0.2) +
  stat_ma_line() +
  stat_ma_eq(use_label(c("eq", "theta", "R2", "P"))) +
  theme_bw()
```

::: callout-note

[*Documentation web site*](http://docs.r4photobiology.info/ggpmisc/) includes all help pages, with output from all examples and vignettes in HTML format .

Please raise issues concerning bugs or enhancements to this package through GitHub at <https://github.com/aphalo/ggpmisc/issues>. Pull requests are also welcome.

:::
