---
title: "R for Photobiology"
subtitle: "Difficult Computations Made Easy"
author: "Pedro J. Aphalo"
date: 2026-04-12
date-modified: 2026-04-12
format: 
  live-revealjs:
    self-contained: true
    mermaid-format: png
    mermaid:
      theme: neutral
    slide-level: 3
self-contained-math: true
engine: knitr
webr:
  packages:
    - knitr
    - lubridate
    - SunCalcMeeus
    - photobiology
    - ggspectra
    - photobiologyWavebands
    - photobiologyPlants
prompt: false
code-fold: show
code-summary: "+/-"
code-tools: true
categories: [slides from talk, R, r4photobiology]
keywords: [R programing, R packages, photobiology]
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

## Introduction

### Why and how did I develop the packages?

- Because I am lazy, but mostly in the long term
  - ... I like efficiency + reliability
- Approach: write code well once and use many times
- Document: for future me and for others
- Test cases: avoid errors now and in the future
- Write clear code: make enhancements and corrections easy

### Did it work out as I expected?

- Expected: In the long run I did save time and effort
- Expected: Became easier to work with large data sets
- Expected: Other researchers are using the packages
- Unexpected: As it made calculations a lot easier
- ... I did "what if" calculations more frequently
- ... leading to new insights and original research
- Unexpected: Citations!

### Why did I use R?

- I had been using R for > 10 years
  - ... as it is the best language for statistics
- The logic of R fits well with how my brain works
  - ... it feels natural to me (not necessarily to you)
- R had the best plotting approach available
  - ... and I was familiar with it
- The package system of R is very well structured

### Side note on R

::: aside
- I think all PhD (and MSc) students of biology and agriculture should learn R
- For many jobs R is considered a big plus
- After graduation, wherever you are you will have R available
  - it is free and open-source
  - works everywhere from mainframes to iPhone, even in web-browsers
- It is very stable
  - new features do not break old code
- It supports almost every existing statistical method
  - reuse what you already know and study just what you need
- **Difficulty:** Thesis supervisors need also to learn R
  - ... and supervisors frequently lack the time
:::

### Design aims for my *3rd* attempt

I made bad design decisions in my first two attempts at writing an R package for photobiology. My third attempt and current packages have as aims:

- Avoid accidental loss of information
- Automate all that can be automated
- Allow customizations
- Store metadata together with the data
- (Fast computations on large data)

### Design aims in detail

- Automate all that can be automated
- Allow customizations
- Store metadata together with the data
- Avoid accidental loss of information

### Design aims in detail

- **Automate all that can be automated**
  - Known units and bases of expression
  - Provide good defaults, but allow overriding them
- Allow customizations
- Store metadata together with the data
- Avoid accidental loss of information

### Design aims in detail

- Automate all what can be automated
- **Allow customizations**
  - Open source code
  - Export "building blocks" to allow extensions
- Store metadata together with the data
- Avoid accidental loss of information

### Design aims in detail

- Automate all that can be automated
- Allow customizations
- **Store metadata together with the data**
  - What was measured
  - When was it measured
  - Where was it measured
  - How was it measured
  - Store a trace of operations applied
- Avoid accidental loss of information

### Design aims in detail

- Automate all that can be automated
- Allow customizations
- Store metadata together with the data
- **Avoid accidental loss of information**
  - Use detector pixel wavelengths
  - Keep history of operations on the data
  - (Support undoing of scaling and normalizations)

### Some recent enhancements

- Accurate day and night lengths and position of the sun back in history and into the future
- Support for long time series of spectra
- Fast acquisition of spectra with Ocean Optics spectrometers
- Faster computations: possible to compute summaries for many 1000s of spectra

### Packages

```{r, echo=TRUE}
library(photobiology)
library(photobiologyWavebands)
library(photobiologyPlants)
library(photobiologyLEDs)
library(ggspectra)
set_theme(theme_bw(14))
```

## Basic concepts

### Light spectra $f(\lambda)$ 

```{mermaid}
%%| label: fig-flowchart
%%| fig-cap: Handling of spectral data. Multiple entry and exit points.
%%| fig-align: center
%%| fig-width: 6.5
flowchart TB
  A[Raw counts] --> CC((×)) --> C(Spectral irradiance)
  B[Calibration] --> CC
  H[Import] ---> C
  C -.-> D(Plots)
  C -.-> E(Irradiances)
  C -.-> F(Ratios)
  C --> G[export]
```

### Spectra: energy vs. photons (= quanta)

- Energy irradiance $E_\lambda$ in $W\,m^{-2}\,nm^{-1}$
- Photon irradiance $Q_\lambda$ in $mol\,s^{-1}\,m^{-2}\,nm^{-1}$
- ... conversion based on energy of one mole of photons
$$q^\prime = h\cdot N_\mathrm{A} \cdot \frac{c}{\lambda}$$
where $\lambda$ is wavelength, and $c$, speed of light, $h$, Planck's constant and $N_\mathrm{A}$ Avogadro's number are constants.
- functions `e2q()` and `q2e()` in R package 'photobiology'

## Plots

```{mermaid}
%%| fig-align: center
%%| fig-width: 6.5
flowchart TB
  C(**Spectral irradiance**)
  C ==> D(**Plots**)
  C -.-> E(Irradiances)
  C -.-> F(Ratios)
style E fill:#fff,color:#aaa,stroke-dasharray: 3 4
style F fill:#fff,color:#aaa,stroke-dasharray: 3 4
```

### Data and functions used for plots

- Spectral data from package 'photobiology'
  - `sun.spct` spectral irradiance
  - `sun_daily.spct` spectral daily exposure
- Methods from package 'ggspectra' ('ggplot2')
  - `autoplot()`

### Sunlight: _energy_

```{r, echo=TRUE}
autoplot(sun.spct, unit.out = "energy")
```

### Sunlight: _photons_

```{r, echo=TRUE}
autoplot(sun.spct, unit.out = "photon")
```

### Spectrum plot: **playground**  {.scrollable}

```{webr}
autoplot(sun.spct, unit.out = "photon")
```

### Sunlight daily: _photons_

```{r, echo=TRUE}
autoplot(sun_daily.spct, unit.out = "photon")
```

## Irradiance

```{mermaid}
%%| fig-align: center
%%| fig-width: 6.5
flowchart TB
  C(**Spectral irradiance**)
  C -.-> D(Plots)
  C ==> E(**Irradiances**)
  C -.-> F(Ratios)
style D fill:#fff,color:#aaa,stroke-dasharray: 3 4
style F fill:#fff,color:#aaa,stroke-dasharray: 3 4
```

### Data and functions used for plots

- Spectral data from package 'photobiology'
  - `sun.spct` spectral irradiance
  - `sun_daily.spct` spectral daily exposure
- Methods from package 'photobioloy'
  - `e_irrad()`
  - `q_irrad()`
  - `waveband()`

### Irradiance: energy vs. photons (= quanta)

- Energy irradiance $E$ in $W\,m^{-2}$
- Photon irradiance $Q$ in $mol\,s^{-1}\,m^{-2}$
- ... no direct conversion possible!
- $\to$ compute $E$ and $Q$ from spectrum

### (Energy) Irradiance $E$

$$E = \int_{\lambda_1}^{\lambda_2} E_\lambda\ d\,\lambda$$
where the wavelength range $\lambda_1$ to $\lambda_1$ is the _waveband_ to integrate.

### (Energy) Irradiance $E$ {.scrollable}

```{r, echo=TRUE}
# VIS in W m-2
e_irrad(sun.spct, waveband(c(400, 500))) # a number with metadata
```

```{r, echo=TRUE}
# VIS in W m-2
as.numeric(e_irrad(sun.spct, waveband(c(400, 500)))) # a plain number
```

### Photon Irradiance (=PFD) $Q$

$$Q = \int_{\lambda_1}^{\lambda_2} Q_\lambda\ d\,\lambda$$
where the wavelength range $\lambda_1$ to $\lambda_1$ is the _waveband_ to integrate.

### Photon Irradiance (=PFD) $Q$ {.scrollable}

```{r, echo=TRUE}
# PAR in mol m-2 s-1
q_irrad(sun.spct, waveband(c(400, 500)))
# PAR in umol m-2 s-1
q_irrad(sun.spct, waveband(c(400, 500)), scale.factor = 1e6)
# PAR in mol m-2 d-1
q_irrad(sun_daily.spct,waveband(c(400, 500)))
```

### Irradiance and PFD: Wavebands {.scrollable}

```{webr}
q_irrad(sun.spct, waveband(c(400, 500)), scale.factor=1e6)
q_irrad(sun.spct, waveband(c(400, 500), wb.name = "BL"),
        scale.factor=1e6)
```

### Wavelength ranges by name {.scrollable}

Multiple definitions in use!

```{r, echo=TRUE}
q_irrad(sun.spct, 
        list(Red("Warrington"), Red("Smith20"),
             Red("Sellaro"), Red("Apogee")), 
        scale.factor = 1e6, return.tb = TRUE)
```

```{r, echo=TRUE}
wl_range(Red("Warrington"))
wl_range(Red("Smith20"))
wl_range(Red("Sellaro"))
wl_range(Red("Apogee"))
```

### Wavelength ranges by name {.scrollable}

```{webr}
Blue()
q_irrad(sun.spct, Blue(), scale.factor=1e6)
```

### Multiple wavelength ranges {.scrollable}

```{webr}
q_irrad(sun_daily.spct, VIS_bands(), scale.factor=1e6)
```

### Spectral weighting functions {.scrollable}

Energy irradiance in 400 to 700 nm is not **PAR**, even if meteorologists call it PAR. I use **PhR** instead.

```{r, echo=TRUE}
# PhR in W m-2
e_irrad(sun.spct, PhR())
# PAR in umol m-2 s-1
q_irrad(sun.spct, PAR(), scale.factor = 1e6)
```

### Spectral weighting functions {.scrollable}

```{webr}
q_irrad(sun.spct, PAR(), scale.factor=1e6)
q_irrad(sun.spct, PAR("ePAR"), scale.factor=1e6)
e_irrad(sun.spct, PG())
```

### Multiple spectra {.scrollable}

```{webr}
summary(sun_evening.spct)
```

```{webr}
q_irrad(sun_evening.spct, Plant_bands(),
        scale.factor = 1e-6)
```

## Ratios

```{mermaid}
%%| fig-align: center
%%| fig-width: 6.5
flowchart TB
  C(**Spectral irradiance**)
  C -.-> D(Plots)
  C -.-> E(**Irradiances**)
  C ==> F(Ratios)
style D fill:#fff,color:#aaa,stroke-dasharray: 3 4
style E fill:#fff,color:#aaa,stroke-dasharray: 3 4
```

### Ratios: energy vs. photons (= quanta)

- Energy ratio $E:E$ $\neq$ photon ratio $Q:Q$
- Unitless
- _shape of the spectrum matters_
- ... no direct conversion possible!
  - compute from spectrum
  - functions `e_ratio()` and `q_ratio()`
  - functions `eq_ratio()` and `qe_ratio()`

### Photon Ratios {.scrollable}

```{r, echo=TRUE}
# R:FR according to Harry Smith's 20 nm definition
R_FR(sun.spct)
# R:FR according to Harry Smith's 20 nm definition
q_ratio(sun.spct, Red("Smith20"), Far_red("Smith20"))
# B:R according to Sellaro's wider definition
q_ratio(sun.spct, Blue("Sellaro"), Red("Sellaro"))
```

```{r, echo=TRUE}
# R:FR according to Harry Smith's narrow definition
q_ratio(sun.spct, Red("Smith10"), Far_red("Smith10"))
# R:FR according to Harry Smith's narrow definition
R_FR(sun.spct, "Smith10")
```

## Phytochrome photoequilibrium

Can be estimated from:
- the wavelength of single colour light
- from the spectrum of the light
- for a mix of red and far red light

### Single wavelength {.scrollable}

```{r, echo=TRUE}
Pfr_Ptot(660)
round(
  Pfr_Ptot(c(435, 460, 530, 630, 660, 710, 735)),
  2)
```

### Spectrum {.scrollable}

```{r, echo=TRUE}
round(Pfr_Ptot(sun.spct), 3)
```

## Daylength and sun position

### Daylength in Beijing {.scrollable}

```{r, echo=TRUE}
CAAS.Beijing <- data.frame(lat = 39.9596, 
                           lon = 116.3188)
day_length(geocode = CAAS.Beijing)
```

```{r, echo=TRUE}
day_length(date = c(as.POSIXct("2025-12-21"),
                    as.POSIXct("2026-06-21")),
           geocode = CAAS.Beijing)
```
### Current position of the sun {.scrollable}

```{webr}
CAAS.Beijing <- data.frame(lat = 39.9596, 
                           lon = 116.3188,
                           address = "CAAS")
sun_angles(time = lubridate::now(tzone = "Asia/Shanghai"),
           geocode = CAAS.Beijing)[, 1:8]
```

## Operations on spectra

- `*`, `+`, etc.
- simulate effect of a filter
- mix light sources
- estimate response

### Sunlight below a yellow filter

```{r, echo=TRUE}
filtered_sun.spct <- sun.spct * yellow_gel.spct
autoplot(filtered_sun.spct, unit.out = "photon")
```

### Two LED spectra

Normalized spectra so that $\mathrm{max} Q_\lambda = 1$.

```{r, echo=TRUE}
autoplot(leds.mspct[c("Nichia_NF2W757GT_F1_sm505_Rfc00",
                      "LedEngin_LZ1_10R302_740nm")],
  unit.out = "photon", norm = "max") +
  theme(legend.position = "none")
```


### Combine two LEDs {.scrollable}

Spectrum with 200 $\mu mol m^{-2} s^{-1}$ from white Nichia Optisolis LED and 20 $\mu mol m^{-2} s^{-1}$ from an Ledengin-OSRAM fra-red LED.

```{r, echo=TRUE}
white_led_200 <-
  fscale(leds.mspct$Nichia_NF2W757GT_F1_sm505_Rfc00, 
         f = q_irrad, 
         target = 200e-6, 
         w.band = PAR())

far_red_20 <-
  fscale(leds.mspct$LedEngin_LZ1_10R302_740nm,
         f = q_irrad, 
         target = 20e-6)

combined.spct <- white_led_200 + far_red_20
```

### Plot computed spectrum

```{r, echo=TRUE}
autoplot(combined.spct, unit.out = "photon")
```

### PAR and ePAR  {.scrollable}

```{r, echo=TRUE}
q_irrad(combined.spct, 
        list(PAR(), PAR("ePAR")),
        scale.factor = 1e6, return.tb = TRUE)
q_irrad(white_led_200,
        list(PAR(), PAR("ePAR")),
        scale.factor = 1e6, return.tb = TRUE)
```

### Red to far-red photon ratio (R:FR)

```{r, echo=TRUE}
R_FR(combined.spct)
R_FR(white_led_200)
```

### Phytochrome photoequilibrium

```{r, echo=TRUE}
Pfr_Ptot(combined.spct)
Pfr_Ptot(white_led_200)
```

## Thanks!

- Thanks for listening today!
- Thanks for looking after me so well while in Beijing!
- Thanks for opening the doors of your group to me!

### Resources

This presentation is a ["web page"](https://www.r4photobiology.info/presentations/r4p-caas-2026.html#/title-slide) and the editable '.qmd' markdown source file is at [github](https://github.com/aphalo/r4photobiology-intro-2026).

::: aside
- My very simple [Introduction to R](https://r4photobiology.info/pages/R-intro.html) as a web page
- A proper [Introduction to R](https://intro2r.com/) from other authors
- My book [Learn R: As a Language](https://www.learnr-book.info/) and a [preprint of the 2024 2nd edition](https://researchportal.helsinki.fi/files/374547141/Aphalo2024-Learn-R-book-2ed-preprint.pdf)
- Each of my R packages has its own [documentation website](https://docs.r4photobiology.info/).
- [Handbook on UV Photobiology Methods](http://doi.org/10.31885/9789521083631)
:::
