---
title: "Spectral Irradiance with 'ooacquire'"
subtitle: "A Brief Tutorial (PARTIAL DRAFT)"
author: "Pedro J. Aphalo"
date: 2023-04-27
date-modified: 2023-04-27
categories: [R Package Update]
keywords: [ooacquire pkg]
code-fold: false
---

::: callout-warning
The description and diagrams shown in this page are valid for 'ooacquire'
version 0.3.2-1. Versions <= 0.3.2 differ from version 0.3.2-1 differ in the
user interface implemented in function `acq_irrad_interactive()` described in
this page.
:::

## Introduction

R package ['ooacquire'](https://docs.r4photobiology.info/ooacquire/) is a
component of the [R for
Photobiology](https://www.r4photobiology.info/pages/r4p-introduction.html) suite
of packages and supports the near-real-time acquisition of spectral data using
Ocean Optics spectrometers from [Ocean Insight](https://www.oceaninsight.com/). 
Package 'ooacquire' imports and uses classes, methods and functions defined in other
packages of the suite.

Package 'ooacquire' has been evolving for more than 10 years and will continue
to evolve, specially with respect to interactive acquisition of spectral data. 
The user interface for spectral irradiance is currently more polished and complex than
the interfaces for fluence spectra and transmittance and refleectance spectra.
This is because both myself and other users most of the time use 'ooacquire'
to measure spectral irradiance. 

The user interfaces (UIs) implemented in functions `acq_irrad_interactive()`,
`acq_fluence_interactive()` and `acq_fraction_interactive()` are text based
rather than graphical. The design is optimized for frequent users rather than
for occasional users. A key design objective has been to make the acquisition of
multiple spectra per measurement session as comfortable and as fast as possible,
i.e., avoid tedious. This is different to what is usually meant by _user
friendly_ involving hand-holding and verbose options within the interface. 

As much as possible, options are selected by typing a single character, or even
just the enter key, thanks to context and state dependent defaults. The
interface is as far as it can be from visually attractive, but it works well for
myself, and most other people who use it.

To start a session within R for the measurement of spectral irradiance
accepting all defaults we can type at the R console, one statement to
load and attach package 'ooacquire', a second, optional statement to set an
option to tell R to print warning immediately instead of delayed, and a
third statement to call the function and then use the UI it opens.

```{r, eval=FALSE}
library(ooacquire)
options(warn = 1)
acq_irrad_interactive()
```

The video below demostrates the use of the interface for an old version of
'ooacquire' (will be replaced by an up-to-date one).

{{< video https://vimeo.com/821595528 >}}

The UI implemented in `acq_irrad_interactive()` is rather _terse_ and, thus,
needs to be clearly documented. The vignettes of package 'ooacquire' include
some examples. However, a description of the logic can be best done using
diagrams. [Mermaid](https://mermaid.js.org/) is a markdown-like language for the
easy creation of diagrams, but not yet supported in vignettes. As its many
diagrams are encoded using Mermaid and it includes video content, this tutorial
is published as a page at the [R for Photobioly web
site](https://www.r4photobiology.info) instead of as part of the 'ooacquire'
R pakage documentation.

## Interface modes

Function `acq_irrad_interactive()` is like a Swiss Army Knife. It has multiple
''personalities'' and the first step is the choice of `interface.mode`, which
determines which interface will be shown to the user. Basically, the selected
`interface.mode` determines which options can be modified by the user through
the interactive UI and which ones are fixed at the values set when the function
is called. 

In addition, whether PDF files containing plots of individual measurements, and
whether different summaries and plots from the created collectiond of spectra
are saved to files or not can be controlled also by passing arguments to formal
parameters when the function is called.

Normally, when measuring several spectra in a row, any redundant need to select
choices in a menue or enter repeatedly the same setting values by the user is
distracting and time consuming. This is tackled using three three strategies: 1)
with multiple interface modes with each mode displaying the parts of the user
interface used in the case of a given type of measurements, and 2) by
context-dependent defaults in menues, and 3) by easily repeating measurements
with the most recently used settings and reference readings.

In this section we describe the overall logic behind strategies 1) and 3). Four
modes are available, and which one is used is selected by the argument passed to
formal parameter `interface.mode` when `acq_irrad_interactive()` is called to
start a spectral-data acquisition session.
There are four main interface modes built into `acq_irrad_interactive()` with 
`"auto"` as default.

```{mermaid}
%%{ init: {"graph": {"htmlLabels": true}} }%%
%%| fig-align: center
flowchart TB
  A[interface.mode = ] -.-> B("simple")
  A --> C(<b>auto</b>)
  A -.-> D("series")
  A -.-> E("full")
```

Appending `-attr` to these mode names enables the UI for setting the values of 
attributes `what.measured` and `comment`. For example, `"simple"` becomes
`"simple-attr"`.

## Start up

The UI does not differ among interface modes for the initialization steps, and
it is not part of the loop for the acquisition of individual spectra or series
of spectra.

```{mermaid}
%%| fig-align: center
flowchart TB
  A((<b>start</b>)) --> B(select spectrometer\nand channel?)
  B --> C(session name?) --> C1(user name?) --> D(folder name?) --> E(((<b>1</b>)))
```

The loop diagrams shown in the sections below, are all active after these first 
steps are completed. The encircled 1, simply signals this connection.

## User Interface Loop

In the diagrams below I show single boxes not only for the input of single values but also for
menus containing multiple options. These menues are `attribute?`, `acquisition parameters?`, 
and `series parameters?`. Within `plot and save` and `plot and save collection`
there are options about the formatting of the plot, and in the second case also
about the file name used to save the collection.

We start with a complete flowchart of the UI loop for the acquisition of 
spectra, which shows all the possibilities, but does not correspond to any
`interface.mode`, because all of them are subsets.

Once one measurement is completed and saved, a new measurement can be acquire
using the same settings for the acquisition as well as reusing the same _dark_ and
_filter_ reference spectra (`repeat`) or with the possibility of changing
these settings and acquiring fresh _dark_ and _filter_ reference spectra (`next`).

```{mermaid}
%%| fig-align: center
flowchart TB
  E(((<b>1</b>))) --> G(protocol?)
  G --> I(object name?) --> J(attributes?) --> K(acquisition\nparameters?)
  K --> L(series parameters?) --> M[[measure\nspectra]] --> N[(plot and save)] 
  N --> N1{collect?} --yes--> N2[(plot and save\ncollection)] --> O{next\n repeat\n or end?}
  N1 --no--> O
  I -.repeat.-> L
  O -.repeat.-> I
  O --next--> G
  O --> P((<b>end</b>))
```

### `"simple"` mode

The `"simple"` interface mode provides a minimalist UI, quite similar to the only UI available
in early versions of 'ooacquire'. It is intended for the acquisition of individual spectra.

```{mermaid}
%%| fig-align: center
%%{ init: {"graph": {"htmlLabels": true}} }%%
flowchart
  E(((<b>1</b>))) --> G(protocol?)
  G --> I(object name?) --> K(acquisition\nparameters?\n<i>simplified</i>)
  K --> M[[measure\nspectra]] --> N[(plot and save)] --> O{next\nrepeat\nor end?}
  I -.repeat.-> M
  O -.repeat.-> I
  O --next--> G
  O --> P((<b>end</b>))
```

### `"auto"` mode

The `"auto"` interface mode provides a complete UI intended for the acquisition of individual spectra and possibly packing groups of spectra into collections.

```{mermaid}
%%| fig-align: center
%%{ init: {"graph": {"htmlLabels": true}} }%%
flowchart TB
  E(((<b>1</b>))) --> G(protocol?)
  G --> I(object name?) --> K(acquisition\nparameters?\n<i>full</i>)
  K --> M[[measure\nspectra]] --> N[(plot and save)]
  N --> N1{collect?} --yes--> N2[(plot and save\ncollection)] --> O{next\nrepeat\nor end?}
  N1 --no--> O
  I -.repeat.-> M
  O -.repeat.-> I
  O --next--> G
  O --> P((<b>end</b>))
```

### `"series"` mode

The `"series"` interface mode provides a complete UI intended for the acquisition of time series of spectra and possibly packing groups of time series of spectra into collections.

```{mermaid}
%%| fig-align: center
%%{ init: {"graph": {"htmlLabels": true}} }%%
flowchart TB
  E(((<b>1</b>))) --> G(protocol?)
  G --> I(object name?) --> K(acquisition\nparameters?\n<i>full</i>)
  K --> L(series parameters?) --> M[[measure\nspectra]] --> N[(plot and save)]
  N --> N1{collect?} --yes--> N2[(plot and save\ncollection)] --> O{next\nrepeat\nor end?}
  N1 --no--> O
  I -.repeat.-> L
  O -.repeat.-> I
  O --next--> G
  O --> P((<b>end</b>))
```

### `"-attr"` variants of the  modes

Appending `"-attr"` to the name of any `interface-mode` enables the menue
and dialogues that enable the setting of the `what.measured` and `comment`
attributes of the `source_spct` objects into which spectra are saved. This
can be represented in the diagrams by the insertion of the `attributes?` box 
between `object name?` and `acquistion parameters?` as shown in the first
diagram showing the full logic. 

### Selecting the interface mode

When calling function `acq_irrad_interactive()` we can pass arguments to
tailor the UI to our current needs. 
For example, to use the `"simple"` mode instead of the default `"auto"` mode
we can call the function as follows.

```{r, eval=FALSE}
acq_irrad_interactive(interface.mode = "simple")
```

We can in addition disable the saving of specific output to files. For example
to disable the saving of PDF files of the plots from individual measurements and
from collections of spectra we can use:

```{r, eval=FALSE}
acq_irrad_interactive(save.pdfs = FALSE)
```

If we do not intend to create and save collections of spectra or summaries, we can
disable these steps, removing from users' view one menu that becomes redundant.

```{r, eval=FALSE}
acq_irrad_interactive(save.collections = FALSE, 
                      save.summaries = FALSE)
```

If we do not want to save ready made summaries but still be able to create
and save collections of spectra we can use.

```{r, eval=FALSE}
acq_irrad_interactive(save.summaries = FALSE)
```

Of course, these options are independent of each other, and can be combined as 
needed.

::: callout-note
Saving plots of large collections of spectra or of long time series of spectra 
to disk is a slow process. Disabling this step can be worthwhile in such cases.
:::

## Measurement protocols

The different diagrams below _open_ the actions that take place within the
`measure spectra` boxes in the diagrams above, previously treated as _black
boxes_. These actions depend on the protocol used for the measurements and on
whether the measurement is a fresh one (`next`) or one reusing settings and
reference measurements (`repeat`). The same measurement protocols are available
in the different user interface modes. We show here the protocols available by
default. However, users can define their own protocols. In the diagrams below,
`start` and `end` represent the start and end of a measurement event, i.e., the
boundaries of the generic `measure spectra` boxes in the diagrams above.

### Protocol `l`

`l` stands for acquisition of _light_ scans.

```{mermaid}
flowchart LR
  S((<b>start</b>)) --> C
  C[light\nspectrum or\nspectra] --> E((<b>end</b>))
```

### Protocol `ld`

`ld` stands for acquisition of _light_ and _dark_ scans sequentially in this order.

```{mermaid}
flowchart LR
  S((<b>start</b>)) --> A
  A -.repeat.-> E
  A[light\nspectrum or\nspectra] --next--> D[dark\nspectrum] --> E((<b>end</b>))
```

### Protocol `lfd`

`lfd` stands for acquisition of _light_, _filter_ and _dark_ scans sequentially in this order.

```{mermaid}
flowchart LR
  S((<b>start</b>)) --> A
  A -.repeat.-> E
  A[light\nspectrum or\nspectra] --next--> B[filter\nspectrum] --> C[dark\nspectrum] --> E((<b>end</b>))
```

### Protocol `dl`

`dl` is identical to `ld` except for the sequential order of the acquisition of data.

```{mermaid}
flowchart LR
  S((<b>start</b>)) --next--> A
  S -.repeat.-> C
  A[dark\nspectrum] --> C[light\nspectrum or\nspectra] --> E((<b>end</b>))
```

### Protocol `dfl`

`dfl` is identical to `lfd` except for the sequential order of the acquisition of data.

```{mermaid}
flowchart LR
  S((<b>start</b>)) --next--> A
  S -.repeat.-> C
  A[dark\nspectrum] --> B[filter\nspectrum] --> C[light\nspectrum or\nspectra] --> E((<b>end</b>))
```

## Acquisition parameters

**COMING SOON**
