---
title: "R Packages: Timeline of Updates"
subtitle: "Packages published in CRAN that I maintain "
author: "Pedro J. Aphalo"
date: "2024-07-13"
date-modified: today
categories: [R packages]
keywords: [R, photobiology, data visualization, computations]
freeze: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 3,
  out.width = "100%"
)

library(lubridate)
library(ggplot2)
library(ggpp)
```


```{r, echo=FALSE, include=FALSE}
# protect against date change while script is running
today  <- lubridate::today() 

# download info from CRAN
# cache the data based on R session and date
if (!exists("all_pkgs.df") || 
    is.null(attr(all_pkgs.df, "download.date")) || # not strictly needed
    attr(all_pkgs.df, "download.date") != today) {
  all_pkgs.df <- tools::CRAN_package_db() |>
    tibble::as_tibble()
  attr(all_pkgs.df, "download.date") <- lubridate::today()
}
dim(all_pkgs.df)

all_pkgs.df |> 
  janitor::clean_names()  |> 
  dplyr::filter(grepl("Aphalo", author)) -> my_pkgs.df
dim(my_pkgs.df)

my_pkgs.df |> 
  dplyr::filter(package != "ggrepel") |>
  dplyr::select(Package = package, Title = title, Version = version, Date = published, doi) -> my_pkgs_condensed.df 
dim(my_pkgs_condensed.df)
```


```{r, echo = FALSE}
# Prepare data for plotting
packages <- my_pkgs_condensed.df[["Package"]]
# we download package history only once per session if possible
if (!exists("pkgs_hist.ls") || 
    length(packages) != length(names(pkgs_hist.ls)) ||
    length(setdiff(packages, names(pkgs_hist.ls)))) {
  pkgs_hist.ls <- list() # will collect one member per package
  for (package in packages) {
    pkgs_hist.ls[[package]] <- 
      packageRank::packageHistory(package, check.package = TRUE)
  }
}
dates.vec <- lubridate::Date() # 
packages.vec <- character()
for (package in packages) {
  temp <- pkgs_hist.ls[[package]]
  dates.vec <- c(dates.vec, temp[["Date"]])
  packages.vec <- c(packages.vec, rep(package, length(temp[["Date"]])))
}
ordering <- order(dates.vec)
dates.vec <- dates.vec[ordering]
packages.vec <- packages.vec[ordering]

earliest.date <- dates.vec[1]
latest.date <- dates.vec[length(dates.vec)] 
```

The sources of the R packages I have published are in public Git repositories at [GitHub}(https://github.com/aphalo). Out of the packages that I have authored and maintain, `r length(packages)` are currently available through [CRAN](https://cran.r-project.org/). The total number of packages submissions to CRAN (mostly updates) as author and maintainer is `r length(dates.vec)` since `r format.Date(earliest.date)`.
The most recent of these updates was published in CRAN on `r format.Date(latest.date)`.

```{r 'cran-packages-plot', echo = FALSE,  fig.width=6, fig.height=5}

cran_releases.df <- 
  data.frame(Package = factor(packages.vec), 
             CRAN.dates = dates.vec, 
             ones = rep(1, length(dates.vec)))

# combined plot
ggplot(cran_releases.df, aes(CRAN.dates, Package, colour = Package)) +
  geom_point(shape = "|", size = 6, alpha = 0.8) +
  scale_x_date(name = "Date of Publication in CRAN",
               date_breaks = "year", date_labels = "%Y") +
  theme_minimal(12) + theme(legend.position = "none")
```

```{r, results='asis', echo = FALSE}
#| label: tbl-packages
#| tbl-cap: List of packages available at CRAN-with the most recently updated one at the top. Package names are linked to their pages in CRAN. Titles are linked to their HTML documentation at this site.

my_pkgs_condensed.df |> 
  dplyr::transmute("Package (CRAN)"  = paste("[", Package, "](", "https://doi.org/", doi, ")", sep = ""),
                "Title (documentation)" = paste("[", Title, "](", "https://docs.r4photobiology.info/", package, "/)", sep = ""),
                 Version = Version,
                Date = Date) |>
  dplyr::arrange(desc(Date)) |>
  knitr::kable()
```

Updates under development are published at [R-Universe](https://aphalo.r-universe.dev/builds) as soon as merged or commited into the main branch in the repositories at GitHub. Two packages that depend on a commercial closed-source driver, but usable with a free runtime of the driver, are published only at R-Universe. Some new, still unstable packages, are also only at R-Universe.

R-Universe profile: [https://aphalo.r-universe.dev](https://aphalo.r-universe.dev).
[![:name status badge](https://aphalo.r-universe.dev/badges/:name)](https://aphalo.r-universe.dev/)
[![:packages status badge](https://aphalo.r-universe.dev/badges/:packages)](https://aphalo.r-universe.dev/packages)

