---
title: "Repository migrated to R-Universe"
subtitle: "Updated installation instructions"
author: "Pedro J. Aphalo"
date: "2024-02-10"
date-modified: "2024-02-10"
categories: [website news, repository news]
---

**I have moved the local CRAN-like repository to [R-Universe](https://r-universe.dev/).**
The address <https://r.r4photobiology.info/> is replaced by <https://aphalo.r-universe.dev/>.

## Installation using package 'photobiology' as example

Installation of the most recent stable version from CRAN:

```{r, eval=FALSE}
install.packages('photobiology')
```

Installation of the current pre-release version from R-Universe CRAN-like 
repository and dependencies from CRAN:

```{r, eval=FALSE}
install.packages('photobiology', 
                 repos = c('https://aphalo.r-universe.dev', 'https://cloud.r-project.org'))
```

## Adding the repository to those used by R by default

If you are using a recent version of R, the statements below can be added to the
`.Rprofile` file for the repository to be always checked when installing or
updating packages, or run before installing or updating the packages I maintain.

```{r, eval=FALSE}
repos <- getOption("repos", default = list("CRAN" = 'https://cloud.r-project.org'))
repos[["r4photobiology"]] <- "https://aphalo.r-universe.dev'"
old.repos <- options(repos = repos)
```

In the second case, the previous setting can be restored if needed with the
statement below once the packages have been installed.

```{r, eval=FALSE}
options(repos = old.repos)
```

